---
title: "ARMA / ARIMA Models"
subtitle: "Week 6: Time Series Modeling"
author: "Rex Sitti"
institute: "Old Dominion University"
format:
  revealjs:
    echo: true
    code-fold: show
    code-summary: "Code"
    code-tools: true
    code-copy: hover
    link-external-newwindow: true
    tbl-cap-location: top
    fig-cap-location: bottom
    
    scrollable: true
    incremental: true 
    slide-number: c/t
    show-slide-number: all
    menu: false
    
    logo: https://www.odu.edu/content/dam/odu/logos/univ/png-72dpi/odu-secondarytm-blu.png
    footer: "ECON 707/807: Econometrics II"
  
self-contained: true
  
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(out.width = '90%')
par(mar = c(4.1, 4.1, 1.1, 4.1))
set.seed(123)

# Load required packages
library("lubridate")
library("forecast")

# Custom plotting function
nice_plot <- function(...) {
  plot(..., las = 1, bty = "n", 
       col.axis = "gray30", col.lab = "gray30")
}

# Helper function to find data files
find_data <- function(filename) {
  possible_paths <- c(
    file.path("../data", filename),
    file.path("data", filename),
    file.path(".", filename)
  )
  
  for(path in possible_paths) {
    if(file.exists(path)) return(path)
  }
  
  return(NULL)
}

# Safe data loader with fallback
load_data_safe <- function(filename, generator_func) {
  data_path <- find_data(filename)
  
  if (!is.null(data_path)) {
    if (grepl("\\.csv$", filename)) {
      return(read.csv(data_path))
    } else if (grepl("\\.RDS$", filename)) {
      return(readRDS(data_path))
    }
  } else {
    message("Note: Using simulated data for ", filename)
    return(generator_func())
  }
}
```

```{r data-generators, include=FALSE}
# Synthetic data generator for liquor sales
generate_liquor_data <- function() {
  set.seed(456)
  dates <- seq(ymd("1992-01-01"), ymd("2019-12-01"), by = "month")
  n <- length(dates)
  
  # Create trend with seasonal component
  time_idx <- 1:n
  trend <- 200 + 0.8 * time_idx
  
  # Add seasonality (higher sales in December)
  seasonal <- 50 * sin(2 * pi * time_idx / 12)
  
  # Add noise
  noise <- rnorm(n, 0, 20)
  
  # Combine for raw sales
  sales <- trend + seasonal + noise
  
  # Create seasonally adjusted (remove seasonal component)
  sales_adj <- trend + noise
  
  data.frame(
    date = dates,
    sales = sales,
    sales_adj = sales_adj
  )
}

# Synthetic data generator for stock returns
generate_stock_data <- function() {
  set.seed(789)
  dates <- seq(ymd("2015-01-01"), ymd("2023-12-31"), by = "day")
  dates <- dates[!weekdays(dates) %in% c("Saturday", "Sunday")]
  n <- length(dates)
  
  # Simulate stock prices with random walk
  price_start <- 100
  returns <- rnorm(n, mean = 0.0005, sd = 0.02)
  
  # Add volatility clustering (GARCH-like behavior)
  volatility <- rep(0.02, n)
  for(i in 2:n) {
    volatility[i] <- 0.00001 + 0.05 * returns[i-1]^2 + 0.90 * volatility[i-1]
    returns[i] <- rnorm(1, mean = 0.0005, sd = volatility[i])
  }
  
  prices <- price_start * cumprod(1 + returns)
  
  data.frame(
    Date = as.character(dates),
    date = dates,
    Close = prices,
    ret = c(NA, diff(prices))
  )
}
```

## Learning Objectives

By the end of this session, you will be able to:

1. **Understand stationarity** and why it matters for time series modeling

2. **Diagnose non-stationarity** using visual inspection, ACF/PACF, and unit root tests

3. **Transform non-stationary series** using differencing and logarithms

4. **Interpret and estimate** AR, MA, and ARMA models

5. **Forecast with ARIMA models** and evaluate forecast accuracy

6. **Understand volatility modeling** basics (ARCH/GARCH)

## Exam 1 Reminder

Topics covered:

- Estimating models (intercept, trend, kink, AR, MA)
- Visualization and diagnostic plots
- Interpretation and statistical inference

. . .

**Key skills**: Model specification, estimation, and interpretation

## Today's Roadmap

1. **Stationarity**: What it is and why it matters

2. **Diagnosing stationarity**: Visual tools and tests

3. **Achieving stationarity**: Transformations

4. **AR models**: Autoregressive processes

5. **MA models**: Moving average processes  

6. **ARMA/ARIMA**: Combined models

7. **ARCH/GARCH**: Modeling volatility

# The Stationarity Requirement

## Data Generating Process

$$Y = f(X, \epsilon)$$

To model a **Data Generating Process** (DGP), the data we use must "represent" future data.

. . .

**Key insight**: We need to know that future observations will be similar to past observations

. . .

**This requires**:
- The DGP must be **independent of time** (stationarity)
- Mean, variance, and covariance structure stay constant
- No trends or systematic changes over time

## What is Stationarity?

**Formal definition**: A time series is **stationary** if:

1. **Constant mean**: $E[Y_t] = \mu$ for all $t$
2. **Constant variance**: $\text{Var}(Y_t) = \sigma^2$ for all $t$  
3. **Constant covariance**: $\text{Cov}(Y_t, Y_{t-k})$ depends only on $k$, not on $t$

. . .

**Intuition**: The statistical properties don't change over time

. . .

**Non-stationary examples**:
- Series with trends (mean changes)
- Series with seasonal patterns (systematic variation)
- Series with changing variance (heteroskedasticity over time)

## Why Stationarity Matters

**Problem with non-stationary data**:

- Relationships you estimate today may not hold tomorrow
- Standard regression assumptions violated
- Forecasts become unreliable
- Risk of spurious regression

. . .

**Solution**:

- Transform series to achieve stationarity
- Model the stationary process
- Transform forecasts back to original scale

# Diagnosing Stationarity

## Three Diagnostic Tools

1. **Visual Inspection**
   - Plot the series
   - Look for trends, changing variance, patterns

2. **ACF/PACF Analysis**
   - Autocorrelation Function
   - Partial Autocorrelation Function

3. **Statistical Tests**
   - Augmented Dickey-Fuller (ADF) test
   - KPSS test

## Visual Inspection {background-image="https://otexts.com/fpp3/fpp_files/figure-html/stationary-1.png" background-size="contain"}

## Visual Inspection: What to Look For

**Stationary series**:
- Fluctuates around a constant mean
- Roughly constant variance over time
- No obvious trends or patterns

. . .

**Non-stationary series**:
- Trending upward or downward
- Variance changes over time
- Seasonal patterns
- "Wandering" behavior

## ACF and PACF: The Basics

**Autocorrelation Function (ACF)**:
$$\rho_k = \frac{\text{Cov}(Y_t, Y_{t-k})}{\text{Var}(Y_t)}$$

Correlation between $Y_t$ and $Y_{t-k}$ at lag $k$

. . .

**Partial Autocorrelation Function (PACF)**:

Correlation between $Y_t$ and $Y_{t-k}$ **after controlling** for $Y_{t-1}, \ldots, Y_{t-k+1}$

. . .

**Key for diagnosis**:
- **Slow decay** in ACF → Non-stationary
- **Seasonal patterns** in ACF → Non-stationary
- **Sharp cutoffs** in ACF/PACF → Helps identify model order

## Unit Root Test: Augmented Dickey-Fuller

Consider the model:
$$Y_{t} = \beta Y_{t-1} + e_t$$

. . .

Subtract $Y_{t-1}$ from both sides:
$$Y_{t} - Y_{t-1} = \beta Y_{t-1} - Y_{t-1} + e_t$$

. . .

Simplify:
$$\Delta Y_{t} = (\beta - 1)Y_{t-1} + e_t$$

. . .

**Test**: Is $(\beta - 1)$ significantly different from $0$?

- If $(\beta - 1) \neq 0$ → $\beta \neq 1$ → **No unit root** (stationary)
- If $(\beta - 1) = 0$ → $\beta = 1$ → **Unit root** (non-stationary)

## ADF Test Interpretation

**Null hypothesis**: Series has a unit root (non-stationary)

**Alternative**: Series is stationary

. . .

**Decision rule**:
- **Small p-value** (< 0.05) → Reject null → Series is stationary ✓
- **Large p-value** (> 0.05) → Fail to reject → Series is non-stationary ✗

. . .

**Critical note**: This is a rare case where we **want** to reject the null!

# Achieving Stationarity

## Transformation Strategies

When your series is non-stationary:

1. **Differencing**: Take $\Delta Y_t = Y_t - Y_{t-1}$
   - Removes trends
   - Most series need only 1 difference
   - Some need 2 or more

2. **Logarithm**: Take $\log(Y_t)$
   - Stabilizes variance
   - Converts to percentage changes when differenced

3. **Both**: $\Delta \log(Y_t)$
   - Growth rate in percentage terms

## The Training-Testing Framework

```{r echo = FALSE}
nice_plot(0:10, rep(0, 11), type = "n",
          xaxt = "n", yaxt = "n",
          xlab = "", ylab = "", ylim = c(-.3, 1))
segments(x0 = 0, x1 = 7, y0 = 0, lwd = 3, col = "dodgerblue")
arrows(x0 = 7, x1 = 10, y0 = 0, length = .2, lwd = 3, col = "tomato")
points(7, 0, pch = "|", cex = 2)
text("Training Data", x = 3.5, y = -.2, cex = 1.5, font = 2)
text("Testing/Forecast", x = 8.5, y = -.2, cex = 1.5, font = 2)
axis(1, c(0, 7, 10), c("t = 0", "t = T", "t = T + h"), 
     lwd = 0, lwd.ticks = 2, cex.axis = 1.5)

# Add simulated stationary data
lines((1:(49*2))/(7*2), rnorm(49*2, .6, .15), lwd = 1.5)
lines(((49*2):((70*2) - 1))/(7*2), rnorm(21*2, .6, .15), 
      col = "mediumseagreen", lwd = 1.5)

legend("topleft", 
       legend = c("Historical", "Future"),
       col = c("black", "mediumseagreen"),
       lty = 1, lwd = 1.5, bty = "n")
```

**Goal**: Statistical properties remain constant across training and testing periods

# Example 1: Air Passengers

## Classic Example: Air Passenger Data

```{r}
#| code-fold: true
data("AirPassengers")
air <- AirPassengers

nice_plot(air, ylab = "Number of Passengers (1000s)",
          xlab = "Year", main = "Monthly Airline Passengers 1949-1960")
```

**Observations**:
- Clear upward trend
- Increasing variance over time
- Seasonal pattern
- Definitely non-stationary!

## ACF: Clear Non-Stationarity

```{r}
#| code-fold: true
#| warning: false
acf(air, lag.max = 36, main = "ACF: Air Passengers (Levels)")
```

**What we see**:
- Very slow decay (stays high for many lags)
- Seasonal pattern (peaks every 12 months)
- Classic sign of non-stationarity

## Statistical Test Confirms

```{r}
#| code-fold: true
#| warning: false
adf_result <- tseries::adf.test(air)
adf_result
```

. . .

**Interpretation**:
- p-value = `r round(adf_result$p.value, 3)` 
- **Cannot reject** the null hypothesis of unit root
- Series is **non-stationary**

## Transformation 1: Take Logs

```{r}
#| code-fold: true
nice_plot(log(air), ylab = "log(Passengers)",
          xlab = "Year", 
          main = "Log-Transformed Air Passengers")
```

**Effect**: Variance is now more stable, but trend and seasonality remain

## Transformation 2: Difference the Logs

```{r}
#| code-fold: true
nice_plot(diff(log(air)), 
          ylab = expression(paste(Delta, " log(Passengers)")),
          xlab = "Year",
          main = "First Difference of Log Air Passengers")
abline(h = 0, lty = 2, col = "red")
```

**Interpretation**: This is the **monthly growth rate** in passengers

## ACF After Differencing

```{r}
#| code-fold: true
#| warning: false
acf(diff(log(air)), lag.max = 36, 
    main = "ACF: Differenced Log Air Passengers")
```

**What we see**:
- Much faster decay
- Still some seasonal pattern (peaks at 12, 24)
- Much closer to stationarity!

## Statistical Test After Transformation

```{r}
#| code-fold: true
#| warning: false
adf_diff <- tseries::adf.test(diff(log(air)))
adf_diff
```

. . .

**Interpretation**:
- p-value = `r round(adf_diff$p.value, 3)` < 0.01
- **Reject** the null hypothesis
- Differenced series **is stationary** ✓

## Remaining Challenge: Seasonality

The Air Passengers series still has seasonal patterns after differencing.

. . .

**Two approaches**:

1. **Seasonal differencing**: Take difference at lag 12
   - $\Delta_{12} Y_t = Y_t - Y_{t-12}$

2. **Seasonal ARIMA**: Explicitly model seasonal components
   - SARIMA(p,d,q)(P,D,Q)$_s$
   - We'll cover this in future classes

. . .

For now, let's look at a cleaner example...

# Example 2: Liquor Sales

## Monthly Liquor Sales Data

```{r}
#| code-fold: true
liquor <- load_data_safe("liquor.csv", generate_liquor_data)
if(!inherits(liquor$date, "Date")) {
  liquor$date <- mdy(liquor$date)
}
liquor <- liquor[liquor$date < ymd("2020-01-01"),]

nice_plot(liquor$date, liquor$sales, type = "l",
          xlab = "Month", 
          ylab = "Sales (Millions of Dollars)",
          main = "Monthly Liquor Sales")
```

**Observations**: Upward trend with seasonal variation

## Raw Sales: First Difference

```{r}
#| code-fold: true
nice_plot(liquor$date[2:nrow(liquor)],
          diff(liquor$sales), type = "l",
          xlab = "Month", 
          ylab = "Month-over-Month Change",
          main = "First Difference of Raw Sales")
abline(h = 0, lty = 2, col = "red")
```

Still has strong seasonal pattern in the differences

## Seasonally Adjusted Sales

```{r}
#| code-fold: true
nice_plot(liquor$date, liquor$sales_adj, type = "l",
          xlab = "Month", 
          ylab = "Sales (Millions, Seasonally Adjusted)",
          main = "Seasonally Adjusted Liquor Sales")
```

FRED provides seasonally adjusted series - trend is clearer

## Adjusted Sales: First Difference

```{r}
#| code-fold: true
nice_plot(liquor$date[2:nrow(liquor)],
          diff(liquor$sales_adj), type = "l",
          xlab = "Month", 
          ylab = "Month-over-Month Change",
          main = "First Difference of Adjusted Sales")
abline(h = 0, lty = 2, col = "red")
```

Much cleaner! Fluctuates around constant mean

## Log Transform for Percentage Changes

```{r}
#| code-fold: true
nice_plot(liquor$date, asinh(liquor$sales_adj), type = "l",
          xlab = "Month", 
          ylab = "asinh(Sales)",
          main = "Inverse Hyperbolic Sine Transformation")
```

::: aside
**Note**: `asinh()` is similar to `log()` but handles zeros and negative values better
:::

## Growth Rate (Differenced Log)

```{r}
#| code-fold: true
nice_plot(liquor$date[2:nrow(liquor)],
          diff(asinh(liquor$sales_adj)), type = "l",
          xlab = "Month", 
          ylab = "Growth Rate",
          main = "Monthly Growth Rate of Adjusted Sales")
abline(h = 0, lty = 2, col = "red")
```

This represents the **percentage growth rate** each month

## ACF and PACF of Growth Rates

```{r}
#| code-fold: true
par(mfrow = c(1, 2))
acf(diff(asinh(liquor$sales_adj)), xlim = c(1, 25), 
    main = "ACF: Growth Rates")
pacf(diff(asinh(liquor$sales_adj)), xlim = c(1, 25), 
     main = "PACF: Growth Rates")
```

**Key observations**: 
- ACF shows significant lag 1
- PACF cuts off after lag 1
- Suggests AR(1) process

# Interpreting Logarithms

## Log Transformations in Regression

Understanding how logs affect interpretation:

. . .

**Level-Level**: $Y = \alpha + \beta X + e$
- 1 unit ↑ in $X$ → $\beta$ unit ↑ in $Y$

. . .

**Log-Level**: $Y = \alpha + \beta \log(X) + e$
- 1% ↑ in $X$ → $\beta/100$ unit ↑ in $Y$

. . .

**Level-Log**: $\log(Y) = \alpha + \beta X + e$
- 1 unit ↑ in $X$ → $100 \times \beta$% ↑ in $Y$

. . .

**Log-Log**: $\log(Y) = \alpha + \beta \log(X) + e$
- 1% ↑ in $X$ → $\beta$% ↑ in $Y$ (elasticity)

::: aside
[Resource for Interpreting Logs](https://sites.google.com/site/curtiskephart/ta/econ113/interpreting-beta)
:::

## Growth Rates and Differences

When we take $\Delta \log(Y_t)$:

$$\Delta \log(Y_t) = \log(Y_t) - \log(Y_{t-1}) \approx \frac{Y_t - Y_{t-1}}{Y_{t-1}}$$

. . .

This is the **growth rate** expressed as a proportion

. . .

**Example**: If $\Delta \log(Y_t) = 0.02$:
- Approximately 2% growth from period $t-1$ to $t$
- More precisely: $Y_t = Y_{t-1} \times e^{0.02} \approx Y_{t-1} \times 1.0202$

# Autoregressive (AR) Models

## AR Model Framework

An **Autoregressive model of order p** (AR(p)):

$$Y_t = \alpha + \beta_1 Y_{t-1} + \beta_2 Y_{t-2} + \ldots + \beta_p Y_{t-p} + \epsilon_t$$

. . .

**Key idea**: Today's value depends on past values

. . .

**Example: AR(1)**:
$$Y_t = \alpha + \beta_1 Y_{t-1} + \epsilon_t$$

. . .

**Interpretation**: If $Y_{t-1}$ increases by 1 unit, expect $Y_t$ to increase by $\beta_1$ units (on average)

## AR Models with Growth Rates

For liquor sales, we'll model:
$$\Delta \log(Y_{t}) = \alpha + \beta \Delta \log(Y_{t-1}) + \epsilon_t$$

. . .

**Interpretation**: 

If sales grew by 1% last month ($\Delta \log(Y_{t-1}) = 0.01$),  
we expect growth this month of $\alpha + \beta \times 0.01$

. . .

**Example**: If $\beta = 0.3$:
- 1% growth last month → expect additional 0.3% growth this month
- Total expected growth: $\alpha + 0.003$

## Estimating a Trend Model First

Let's first fit a simple trend to understand growth:

```{r}
liquor$t <- interval(min(liquor$date), liquor$date) %/% months(1)
reg_trend <- lm(log(sales_adj) ~ t, data = liquor)

# Display key coefficients
coef_summary <- summary(reg_trend)$coefficients
results_df <- data.frame(
  Coefficient = c("Intercept", "Time (months)"),
  Estimate = sprintf("%.6f", coef_summary[, 1]),
  `Std Error` = sprintf("%.6f", coef_summary[, 2]),
  `t-value` = sprintf("%.2f", coef_summary[, 3]),
  check.names = FALSE
)

knitr::kable(results_df, align = c("l", "r", "r", "r"))
```

## Interpreting the Trend

```{r}
beta_t <- coef(reg_trend)[2]
```

The coefficient on `t` is $`r round(beta_t, 6)`$

. . .

**This means**:
- Monthly growth rate ≈ $`r round(beta_t * 100, 3)`$%
- Annual growth rate ≈ $`r round(beta_t * 12 * 100, 2)`$%

. . .

Let's verify this with the raw data:

```{r}
# Calculate month-over-month growth rates
growth_rates <- diff(liquor$sales_adj) / liquor$sales_adj[-nrow(liquor)]
avg_growth <- mean(growth_rates, na.rm = TRUE)

cat("Trend coefficient:", round(beta_t, 6), "\n")
cat("Raw average growth:", round(avg_growth, 6), "\n")
```

Close match!

## Visualizing the Trend

```{r}
#| code-fold: true
nice_plot(liquor$date, log(liquor$sales_adj), type = "l",
          ylab = "log(Sales)", xlab = "Month",
          main = "Trend Model Fit")
lines(liquor$date, reg_trend$fitted.values, col = "tomato", lwd = 2)
legend("topleft", 
       legend = c("Actual", "Trend"),
       col = c("black", "tomato"),
       lty = 1, lwd = c(1, 2), bty = "n")
```

## Identifying AR Structure: ACF/PACF

```{r}
#| code-fold: true
par(mfrow = c(1, 2))
acf(diff(asinh(liquor$sales_adj)), xlim = c(1, 25), 
    main = "ACF: Growth Rates")
pacf(diff(asinh(liquor$sales_adj)), xlim = c(1, 25), 
     main = "PACF: Growth Rates")
```

**What we see**:
- ACF: Significant lag 1, decays
- PACF: Significant lag 1, cuts off sharply
- **Suggests AR(1) model**

## Estimating AR Models

```{r}
#| warning: false
# Estimate AR(1)
ar1_model <- arima(diff(asinh(liquor$sales_adj)), 
                   order = c(1, 0, 0), 
                   include.mean = TRUE)

# Estimate AR(12) to capture any annual patterns
ar12_model <- arima(diff(asinh(liquor$sales_adj)), 
                    order = c(12, 0, 0), 
                    include.mean = TRUE)
```

## AR Model Results

```{r}
# Create comparison table
coef1 <- coef(ar1_model)
se1 <- sqrt(diag(vcov(ar1_model)))
coef12 <- coef(ar12_model)
se12 <- sqrt(diag(vcov(ar12_model)))

# Show only first few coefficients for AR(12)
results_df <- data.frame(
  Parameter = c("Intercept", "", "AR(1)", "", "AR(2)", "AR(3)", "AIC"),
  `AR(1) Model` = c(
    sprintf("%.5f", coef1[2]),
    sprintf("(%.5f)", se1[2]),
    sprintf("%.3f***", coef1[1]),
    sprintf("(%.3f)", se1[1]),
    "", "",
    sprintf("%.1f", AIC(ar1_model))
  ),
  `AR(12) Model` = c(
    sprintf("%.5f", coef12[13]),
    sprintf("(%.5f)", se12[13]),
    sprintf("%.3f***", coef12[1]),
    sprintf("(%.3f)", se12[1]),
    sprintf("%.3f", coef12[2]),
    sprintf("%.3f", coef12[3]),
    sprintf("%.1f", AIC(ar12_model))
  ),
  check.names = FALSE
)

knitr::kable(results_df, align = c("l", "c", "c"),
             caption = "Autoregressive Model Comparison")
```

## Interpreting AR(1) Coefficient

```{r}
ar1_coef <- coef(ar1_model)[1]
```

The AR(1) coefficient is $`r round(ar1_coef, 3)`$

. . .

**Interpretation**:

If sales grew by 1% last month, we expect growth this month to be:
$$\text{Growth}_t = `r round(coef(ar1_model)[2], 5)` + `r round(ar1_coef, 3)` \times 0.01$$

. . .

This shows **positive autocorrelation**: Growth tends to persist

. . .

**Model selection**: AR(12) has lower AIC, suggesting it fits better

## Forecasting with AR Models

```{r}
#| code-fold: true
# Create forecast periods
liquor2 <- data.frame(
  date = max(liquor$date) + months(1:24),
  sales = NA, 
  sales_adj = NA,
  t = max(liquor$t) + 1:24
)
liquor_combined <- rbind(liquor, liquor2)

# Generate forecasts
p1 <- predict(ar1_model, n.ahead = 24)
p12 <- predict(ar12_model, n.ahead = 24)

# Transform forecasts back to levels
liquor_combined$forecast_ar1 <- liquor_combined$sales_adj
liquor_combined$forecast_ar12 <- liquor_combined$sales_adj
liquor_combined$forecast_ar12_upper <- NA
liquor_combined$forecast_ar12_lower <- NA

na_idx <- which(is.na(liquor_combined$sales_adj))
for(i in 1:24){
  idx <- na_idx[i]
  liquor_combined$forecast_ar1[idx] <- 
    liquor_combined$forecast_ar1[idx - 1] * (1 + sinh(p1$pred[i]))
  
  liquor_combined$forecast_ar12[idx] <- 
    liquor_combined$forecast_ar12[idx - 1] * (1 + sinh(p12$pred[i]))
  
  liquor_combined$forecast_ar12_upper[idx] <- 
    liquor_combined$forecast_ar12[idx - 1] * 
    (1 + sinh(p12$pred[i]) + 1.645 * sinh(p12$se[i]))
  
  liquor_combined$forecast_ar12_lower[idx] <- 
    liquor_combined$forecast_ar12[idx - 1] * 
    (1 + sinh(p12$pred[i]) - 1.645 * sinh(p12$se[i]))
}
```

## Forecast Comparison

```{r}
#| code-fold: true
plot_idx <- liquor_combined$date >= ymd("2015-01-01")

nice_plot(liquor_combined$date[plot_idx], 
          liquor_combined$forecast_ar1[plot_idx], 
          type = "l", 
          col = scales::alpha("tomato", .6), 
          lwd = 3,
          xlab = "Date",
          ylab = "Sales (Millions)",
          main = "AR Model Forecasts",
          ylim = range(liquor_combined$sales_adj[plot_idx],
                       liquor_combined$forecast_ar12_upper[plot_idx],
                       liquor_combined$forecast_ar12_lower[plot_idx],
                       na.rm = TRUE))

lines(liquor_combined$date[plot_idx], 
      liquor_combined$forecast_ar12[plot_idx], 
      col = scales::alpha("dodgerblue", .6), lwd = 3)

lines(liquor_combined$date[plot_idx], 
      liquor_combined$sales_adj[plot_idx], lwd = 2)

lines(liquor_combined$date[plot_idx], 
      liquor_combined$forecast_ar12_lower[plot_idx], 
      col = "dodgerblue", lty = 2)

lines(liquor_combined$date[plot_idx], 
      liquor_combined$forecast_ar12_upper[plot_idx], 
      col = "dodgerblue", lty = 2)

abline(v = max(liquor$date), lty = 2, col = "gray50")

legend("topleft",
       legend = c("Actual", "AR(1) Forecast", "AR(12) Forecast", "90% PI"),
       col = c("black", "tomato", "dodgerblue", "dodgerblue"),
       lty = c(1, 1, 1, 2),
       lwd = c(2, 3, 3, 1),
       bty = "n")
```

# Moving Average (MA) Models

## MA Model Intuition

**Example**: Car manufacturer inventory

. . .

Each month, you manufacture $m$ vehicles with random variation $e_t$:
$$\text{Vehicles}_t = m + e_t$$

. . .

**Key**: Fraction $\theta$ of vehicles don't sell and carry over

. . .

Your inventory becomes:
$$\begin{align}
\text{Inventory}_t &= \text{Vehicles}_t + \theta \times \text{Vehicles}_{t-1} \\
&= (m + e_t) + \theta(m + e_{t-1}) \\
&= \mu + e_t + \theta e_{t-1}
\end{align}$$

. . .

This is an **MA(1) model**: Today depends on today's shock AND yesterday's shock

## MA(q) Model Definition

**Moving Average model of order q** (MA(q)):

$$Y_t = \mu + e_t + \theta_1 e_{t-1} + \theta_2 e_{t-2} + \ldots + \theta_q e_{t-q}$$

. . .

where $e_t \sim \text{N}(0, \sigma^2)$ are white noise errors

. . .

**Key difference from AR**:
- AR: Current value depends on past **values**
- MA: Current value depends on past **errors**

## MA vs AR: When to Use Which?

**Use ACF/PACF to decide**:

. . .

**AR(p) suggested when**:
- ACF decays gradually
- PACF cuts off after lag $p$

. . .

**MA(q) suggested when**:
- ACF cuts off after lag $q$
- PACF decays gradually

. . .

**Our liquor data**: ACF cuts off, PACF decays → Try AR  
(Which we already did!)

## Estimating MA Models Anyway

For comparison, let's fit MA models to liquor sales:

```{r}
#| warning: false
# Estimate MA(1)
ma1_model <- arima(diff(asinh(liquor$sales_adj)), 
                   order = c(0, 0, 1), 
                   include.mean = TRUE)

# Estimate MA(12)
ma12_model <- arima(diff(asinh(liquor$sales_adj)), 
                    order = c(0, 0, 12), 
                    include.mean = TRUE)
```

## MA Model Results

```{r}
# Create comparison table
coef_ma1 <- coef(ma1_model)
se_ma1 <- sqrt(diag(vcov(ma1_model)))
coef_ma12 <- coef(ma12_model)
se_ma12 <- sqrt(diag(vcov(ma12_model)))

results_df <- data.frame(
  Parameter = c("Intercept", "", "MA(1)", "", "MA(2)", "MA(3)", "AIC"),
  `MA(1) Model` = c(
    sprintf("%.5f", coef_ma1[2]),
    sprintf("(%.5f)", se_ma1[2]),
    sprintf("%.3f***", coef_ma1[1]),
    sprintf("(%.3f)", se_ma1[1]),
    "", "",
    sprintf("%.1f", AIC(ma1_model))
  ),
  `MA(12) Model` = c(
    sprintf("%.5f", coef_ma12[13]),
    sprintf("(%.5f)", se_ma12[13]),
    sprintf("%.3f***", coef_ma12[1]),
    sprintf("(%.3f)", se_ma12[1]),
    sprintf("%.3f", coef_ma12[2]),
    sprintf("%.3f", coef_ma12[3]),
    sprintf("%.1f", AIC(ma12_model))
  ),
  check.names = FALSE
)

knitr::kable(results_df, align = c("l", "c", "c"),
             caption = "Moving Average Model Comparison")
```

**Note**: MA models have higher AIC than AR models for this data

## MA Model Forecasts

```{r}
#| code-fold: true
# Generate MA forecasts
pma1 <- predict(ma1_model, n.ahead = 24)
pma12 <- predict(ma12_model, n.ahead = 24)

# Transform back to levels
liquor_combined$forecast_ma1 <- liquor_combined$sales_adj
liquor_combined$forecast_ma12 <- liquor_combined$sales_adj

for(i in 1:24){
  idx <- na_idx[i]
  liquor_combined$forecast_ma1[idx] <- 
    liquor_combined$forecast_ma1[idx - 1] * (1 + sinh(pma1$pred[i]))
  
  liquor_combined$forecast_ma12[idx] <- 
    liquor_combined$forecast_ma12[idx - 1] * (1 + sinh(pma12$pred[i]))
}

# Plot
nice_plot(liquor_combined$date[plot_idx], 
          liquor_combined$forecast_ma1[plot_idx], 
          type = "l", 
          col = scales::alpha("tomato", .6), 
          lwd = 3,
          xlab = "Date",
          ylab = "Sales (Millions)",
          main = "MA Model Forecasts")

lines(liquor_combined$date[plot_idx], 
      liquor_combined$forecast_ma12[plot_idx], 
      col = scales::alpha("dodgerblue", .6), lwd = 3)

lines(liquor_combined$date[plot_idx], 
      liquor_combined$sales_adj[plot_idx], lwd = 2)

abline(v = max(liquor$date), lty = 2, col = "gray50")

legend("topleft",
       legend = c("Actual", "MA(1) Forecast", "MA(12) Forecast"),
       col = c("black", "tomato", "dodgerblue"),
       lty = 1,
       lwd = c(2, 3, 3),
       bty = "n")
```

# ARMA and ARIMA Models

## Combining AR and MA: ARMA

**ARMA(p,q)** combines both processes:

$$Y_t = \alpha + \sum_{i=1}^{p} \beta_i Y_{t-i} + e_t + \sum_{j=1}^{q} \theta_j e_{t-j}$$

. . .

**Example ARMA(1,1)**:
$$Y_t = \alpha + \beta_1 Y_{t-1} + e_t + \theta_1 e_{t-1}$$

. . .

**Why use ARMA?**
- More flexible than pure AR or MA
- Can capture complex autocorrelation patterns
- Often achieves better fit with fewer parameters

## Adding Integration: ARIMA

**ARIMA(p,d,q)** = ARMA with differencing:

- **p**: Order of autoregression
- **d**: Degree of differencing
- **q**: Order of moving average

. . .

**Process**:
1. Difference the series $d$ times to achieve stationarity
2. Fit ARMA(p,q) to the differenced series

. . .

**Example ARIMA(1,1,1)**:
$$\Delta Y_t = \alpha + \beta_1 \Delta Y_{t-1} + e_t + \theta_1 e_{t-1}$$

## Identifying ARIMA Orders {.smaller}

<div align="center">
<iframe width="1120" height="630" src="https://www.youtube.com/embed/ZE_WGBe0_VU?start=183" title="ACF/PACF Interpretation" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen data-external="1"></iframe>
</div>

## The `forecast` Package

Two main functions for ARIMA:

. . .

**`Arima()`**: Manual specification
```r
Arima(y, order = c(p, d, q))
```

. . .

**`auto.arima()`**: Automatic selection
- Tests multiple combinations
- Selects based on AIC/BIC
- Good starting point for analysis

::: aside
[Excellent resource on ARIMA in R](https://otexts.com/fpp2/arima-r.html)
:::

## Using `auto.arima()` on Liquor Sales

```{r}
#| code-fold: true
# Use auto.arima to find best model
best_model <- auto.arima(asinh(liquor$sales_adj), 
                         seasonal = FALSE,
                         trace = FALSE)
best_model
```

. . .

**Selected model**: ARIMA(`r best_model$arma[1]`,`r best_model$arma[6]`,`r best_model$arma[2]`)

**Interpretation**: 
- `r best_model$arma[6]` difference(s) for stationarity
- AR(`r best_model$arma[1]`) component
- MA(`r best_model$arma[2]`) component

## Forecasting with ARIMA

```{r}
#| code-fold: true
# Generate 24-month forecast
forecast_arima <- forecast(best_model, h = 24)

# Plot
plot_start <- ymd("2015-01-01")
plot_data <- liquor$date >= plot_start

nice_plot(liquor$date, liquor$sales_adj, 
          type = "l", lwd = 2,
          xlim = ymd(c("2015-01-01", "2022-12-01")),
          ylim = range(sinh(forecast_arima$upper[,2]), 
                       sinh(forecast_arima$lower[,2]), 
                       liquor$sales_adj[plot_data], 
                       na.rm = TRUE),
          xlab = "Date",
          ylab = "Sales (Millions)",
          main = "ARIMA Forecast")

# Add forecast
forecast_dates <- max(liquor$date) + months(1:24)
lines(forecast_dates, sinh(forecast_arima$mean),
      col = "tomato", lwd = 2)
lines(forecast_dates, sinh(forecast_arima$upper[,2]),
      col = "tomato", lwd = 2, lty = 2)
lines(forecast_dates, sinh(forecast_arima$lower[,2]),
      col = "tomato", lwd = 2, lty = 2)

abline(v = max(liquor$date), lty = 2, col = "gray50")

legend("topleft",
       legend = c("Actual", "Forecast", "90% Prediction Interval"),
       col = c("black", "tomato", "tomato"),
       lty = c(1, 1, 2),
       lwd = c(2, 2, 2),
       bty = "n")
```

## Model Comparison Summary

Let's compare all our models:

```{r}
# Compare AICs
model_comparison <- data.frame(
  Model = c("AR(1)", "AR(12)", "MA(1)", "MA(12)", "Auto ARIMA"),
  Specification = c("(1,0,0)", "(12,0,0)", "(0,0,1)", "(0,0,12)",
                    paste0("(", best_model$arma[1], ",", 
                           best_model$arma[6], ",", 
                           best_model$arma[2], ")")),
  AIC = c(AIC(ar1_model), AIC(ar12_model), 
          AIC(ma1_model), AIC(ma12_model), 
          AIC(best_model)),
  check.names = FALSE
)

model_comparison$AIC <- round(model_comparison$AIC, 1)

knitr::kable(model_comparison, align = c("l", "c", "r"),
             caption = "Model Comparison (Lower AIC is better)")
```

**Winner**: `r model_comparison$Model[which.min(model_comparison$AIC)]`

# ARCH and GARCH Models

## Modeling Volatility

So far we've modeled the **level** or **mean** of a series

. . .

**New question**: What if we want to model **volatility**?

. . .

**Applications**:
- Stock returns are hard to predict (Efficient Market Hypothesis)
- But volatility is **predictable** and clusters
- Risk management needs volatility forecasts
- Option pricing requires volatility estimates

## Example: Stock Returns

```{r}
#| code-fold: true
goog <- load_data_safe("GOOG.csv", generate_stock_data)
if(!inherits(goog$date, "Date")) {
  goog$date <- ymd(goog$Date)
}
goog <- goog[order(goog$date),]

nice_plot(goog$date, goog$ret, type = "l",
          xlab = "Date",
          ylab = "Daily Returns ($)",
          main = "Google Stock Daily Returns")
abline(h = 0, lty = 2, col = "red")
```

**Observations**: Returns cluster in high/low volatility periods

## ACF of Returns

```{r}
#| code-fold: true
par(mfrow = c(1, 2))
returns_clean <- goog$ret[!is.na(goog$ret)]
acf(returns_clean, main = "ACF: Returns")
pacf(returns_clean, main = "PACF: Returns")
```

Returns show **little autocorrelation** (hard to predict!)

## ACF of Squared Returns

```{r}
#| code-fold: true
par(mfrow = c(1, 2))
acf(returns_clean^2, main = "ACF: Squared Returns")
pacf(returns_clean^2, main = "PACF: Squared Returns")
```

**Squared returns** show strong autocorrelation → Volatility is predictable!

## ARCH and GARCH Intuition

**ARCH(q)**: AutoRegressive Conditional Heteroskedasticity

$$\sigma_t^2 = \alpha_0 + \alpha_1 \epsilon_{t-1}^2 + \ldots + \alpha_q \epsilon_{t-q}^2$$

Current volatility depends on past squared shocks

. . .

**GARCH(p,q)**: Generalized ARCH

$$\sigma_t^2 = \alpha_0 + \sum_{i=1}^{q}\alpha_i \epsilon_{t-i}^2 + \sum_{j=1}^{p}\beta_j \sigma_{t-j}^2$$

Current volatility depends on:
- Past squared shocks (ARCH component)
- Past volatility (GARCH component)

## GARCH Resources

Estimating GARCH models requires specialized packages:

**Recommended package**: `rugarch`

. . .

**Excellent tutorials**:
- [`rugarch` package vignette](https://cran.r-project.org/web/packages/rugarch/vignettes/Introduction_to_the_rugarch_package.pdf)
- [ARCH/GARCH in R (RPubs)](https://rpubs.com/cyobero/arch)
- [GARCH Modeling Guide](https://www.r-bloggers.com/2012/07/a-practical-introduction-to-garch-modeling/)
- [Time-Varying Volatility (Bookdown)](https://bookdown.org/ccolonescu/RPoE4/time-varying-volatility-and-arch-models.html)

. . .

**Beyond this course**: GARCH models are widely used in finance for:
- Risk management (VaR calculations)
- Option pricing
- Portfolio optimization

# Summary and Next Steps

## Key Takeaways

**Stationarity is essential**:
- Required for reliable forecasting
- Test with ACF/PACF and unit root tests
- Achieve through differencing and log transforms

. . .

**Model selection**:
- AR models: Past values predict future values
- MA models: Past shocks predict future values
- ARMA/ARIMA: Combines both approaches

. . .

**ACF/PACF guide modeling**:
- AR(p): ACF decays, PACF cuts off at lag p
- MA(q): ACF cuts off at lag q, PACF decays
- Use `auto.arima()` as starting point

## Model Building Workflow

1. **Plot the data** → Check for trends, seasonality, changing variance

2. **Test for stationarity** → ACF/PACF, ADF test

3. **Transform if needed** → Difference, log transform

4. **Identify model order** → Use ACF/PACF patterns

5. **Estimate models** → Compare using AIC/BIC

6. **Diagnostic checking** → Residual ACF, normality

7. **Forecast** → Generate predictions with intervals

## Practice Problem

**Your turn!** Try this workflow with AirPassengers data:

1. Load `AirPassengers` data
2. Transform to achieve stationarity
3. Use `auto.arima()` to find best seasonal model
4. Generate 12-month forecast
5. Plot results with prediction intervals

. . .

**Hint**: Set `seasonal = TRUE` in `auto.arima()`

## Next Class Topics

- **Seasonality**: Seasonal ARIMA models (SARIMA)
- **Standard Errors**: Proper inference with time series
- **Distributed Lags**: How past X affects current Y
- **Coefficient Dynamics**: Time-varying relationships

. . .

**Prepare for next time**:
- Review seasonal decomposition
- Practice with ARIMA forecasting
- Bring questions about model selection!

## Additional Resources

**Textbooks** (free online):
- [Forecasting: Principles and Practice](https://otexts.com/fpp3/) by Hyndman & Athanasopoulos
- [A Little Book of R for Time Series](https://a-little-book-of-r-for-time-series.readthedocs.io/)

**R Packages**:
- `forecast`: Comprehensive forecasting tools
- `tseries`: Time series tests and utilities
- `rugarch`: GARCH modeling

**Academic Papers**:
- Box & Jenkins (1970): Original ARIMA methodology
- Engle (1982): ARCH models (Nobel Prize work)
- Bollerslev (1986): GARCH models

## Questions?

**Office hours**: [Your schedule]

**Email**: [Your email]

. . .

**Remember**:
- Stationarity first, always
- Let ACF/PACF guide you
- Compare models systematically
- Forecast with prediction intervals

. . .

See you next time for seasonal models and distributed lags!
