
---
title: "Introduction to Time Series Analysis"
subtitle: "ECON 707/807: Econometrics II"
author: "Rex Sitti"
institute: "Old Dominion University - Strome College of Business"
format:
  revealjs:
    slide-number: true
    incremental: true
    logo: https://www.odu.edu/content/dam/odu/logos/univ/png-72dpi/odu-secondarytm-blu.png
    footer: "ECON 707/807: Econometrics II - Spring 2026"
    theme: default
    transition: slide
    background-transition: fade
  pdf:
    documentclass: beamer
execute:
  echo: false
  warning: false
  message: false
code-fold: show
code-summary: "Show Code"
code-tools: true
code-copy: hover
link-external-newwindow: true
tbl-cap-location: top
fig-cap-location: bottom
self-contained: true
editor: source
---

    

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
```

## Time Series

. . .

Sequential measurements or values of a single entity over time.

. . .

Why study time series data?

-   Gross Domestic Product
-   Unemployment
-   Vehicle Demand
-   Energy Consumption

## Forecasting

. . .

"Predict" outcomes of a time series in future (unobserved) periods.

```{r echo = FALSE}
#| output-location: fragment
library("lubridate")
urate <- read.csv("C:/Users/rsitti/OneDrive - Old Dominion University/ODU Classes/Econ 707/data/unrate_us.csv")
colnames(urate) <- c("date", "urate_t")
urate$date <- ymd(urate$date)
plot(urate$date, urate$urate_t,
     ylab = "Unemployment Rate",
     xlab = "Time",
     main = "Monthly Unemployment Rate",
     type = "l", col = "dodgerblue")
```

## Time Series Models {.smaller}

$$Y_{t} = f(Y_{t-1}, X_{t}, X_{t-1}, \tau_t, C_t, S_t)$$ where:

::: nonincremental
-   $Y_{t-1}$ is a lagged value of $Y$
-   $X_t$ is a contemporaneous independent value
-   $X_{t-1}$ is a lagged value of $X$
-   $\tau$ is the trend
-   $C$ is the cycle
-   $S$ is the season
:::

::: aside
Note: this model only considers what is *inside* the model!
:::

## Statistics Review

-   mean
-   variance
-   standard error
-   central limit theorem (CLT)
-   p-value
-   $Y_i = \alpha + \beta X_i + \epsilon_{i}$
-   causality
-   [spurious](https://www.tylervigen.com/spurious-correlations)

## Forecasting Steps

1.  Define Problem

    What am I trying to solve?

2.  Gather Data

    FRED, WRDS, Etc.

3.  Exploratory Data Analysis (EDA)

    Plot, Plot, Plot

4.  Choose & Fit a Model

    $Y_{t} = f(Y_{t-1}, X_{t}, X_{t-1}, \tau_t, C_t, S_t)$

5.  Evaluate, Forecast

## Packages and Functions

-   `table()`
-   `aggregate()`
-   `match()`
-   `paste()`, `paste0()`, `substr()`, `grepl()`, `gsub()`, `regexpr()`, `strsplit()`, `unlist()`

::: aside
I have some materials [here](https://alexcardazzi.github.io/using-r/).
:::

## Packages and Functions {visibility="uncounted"}

-   `for()`
-   `if()`, `else`
-   `ifelse()`
-   `fixest`
-   `scales::alpha()`
-   `data.table::fread()`
-   `readRDS()`, `saveRDS()`

## Packages and Functions {visibility="uncounted"}

-   `lubridate`[^intro_01_new-1]
-   `tsibble`
-   `forecast`
-   `ts()`[^intro_01_new-2]
-   `duplicated()`

[^intro_01_new-1]: [lubridate](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_lubridate.pdf)

[^intro_01_new-2]: [ts](https://www.statmethods.net/advstats/timeseries.html)

## Install and Load Packages

```{r eval = FALSE}
install.packages("lubridate")
library("lubridate")
```

## Make a Time Series

```{r}
#| output-location: fragment
urate <- read.csv("C:/Users/rsitti/OneDrive - Old Dominion University/ODU Classes/Econ 707/data/unrate_us.csv")
colnames(urate) <- c("date", "urate_t")
head(urate)
```

## Plot a Time Series

```{r}
#| output-location: fragment
urate$date <- ymd(urate$date)
plot(urate$date, urate$urate_t,
     xlab = "Month",
     ylab = "Unemployment Rate",
     main = "Monthly Unemployment Rate")
```

## Plot a Time Series {visibility="uncounted"}

```{r}
#| output-location: fragment
plot(urate$date, urate$urate_t,
     ylab = "Unemployment Rate",
     xlab = "Time",
     main = "Monthly Unemployment Rate",
     type = "l", col = "dodgerblue")
```

## Lags (Leads)

```{r echo=FALSE}
head(urate)
```

. . .

```{r echo = TRUE}
#| code-fold: true
t <- 2:nrow(urate)
urate$urate_tm1 <- NA
urate$urate_tm1[t] <- urate$urate_t[t-1]
urate$urate_tp1 <- NA
urate$urate_tp1[t-1] <- urate$urate_t[t]
head(urate)
```

## What about panels?

```{r}
#| code-fold: true
df <- data.frame(state = c(rep("NY", 10),
                           rep("VA", 10),
                           rep("CA", 10)),
                 year = rep(2010:2019, 3),
                 var = rnorm(30))
df[df$year %in% 2010:2012,]
```

## What about panels? {visibility="uncounted"}

```{r}
df$var_lag1 <- NA
```

```{r echo = FALSE}
#| output-location: fragment
df[df$year %in% 2010:2012,]
```

::: aside
Note: these data are already ordered.
:::

## What about panels? {visibility="uncounted"}

```{r}
#| code-fold: true
#| output-location: fragment
df$var_lag1 <- NA
t <- which(duplicated(df$state))
df$var_lag1[t] <- df$var[t-1]
df[df$year %in% 2010:2012,]
```

## Multiple Plots

```{r}
#| output-location: fragment
plot(df$year, df$var)
```

## Multiple Plots {visibility="uncounted"}

```{r}
#| output-location: fragment
plot(df$year, df$var,
     col = match(df$state, unique(df$state)))
```

## Multiple Plots {visibility="uncounted"}

```{r}
#| output-location: fragment
plot(df$year, df$var,
     col = match(df$state, unique(df$state)),
     type = "l")
```

## Multiple Plots {visibility="uncounted"}

```{r}
#| output-location: fragment
lim <- df$state == "NY"
plot(df$year[lim], df$var[lim],
     col = match(df$state[lim], unique(df$state)),
     type = "l")
```

## Multiple Plots {visibility="uncounted"}

```{r}
#| output-location: fragment
lim <- df$state == "NY"
plot(df$year[lim], df$var[lim],
     col = match(df$state[lim], unique(df$state)),
     type = "l")
lim <- df$state == "VA"
lines(df$year[lim], df$var[lim],
     col = match(df$state[lim], unique(df$state)))
lim <- df$state == "CA"
lines(df$year[lim], df$var[lim],
     col = match(df$state[lim], unique(df$state)))
```

## Multiple Plots {visibility="uncounted"}

```{r}
#| output-location: fragment
plot(df$year, df$var, type = "n")
lim <- df$state == "NY"
lines(df$year[lim], df$var[lim],
     col = match(df$state[lim], unique(df$state)))
lim <- df$state == "VA"
lines(df$year[lim], df$var[lim],
     col = match(df$state[lim], unique(df$state)))
lim <- df$state == "CA"
lines(df$year[lim], df$var[lim],
     col = match(df$state[lim], unique(df$state)))
```

## Next Class

1.  $E[y]$, $\hat{y}$
2.  Loss Functions
3.  Lags, Leads
4.  Conditional Forecasts
5.  Time Series Components

## Practice - Cleaning Data

::: nonincremental
1.  Work on [#5](https://foresightbi.com.ng/microsoft-power-bi/dirty-data-samples-to-practice-on/#5_Jumbled_Customer_Details).

2.  Work on [#8](https://foresightbi.com.ng/microsoft-power-bi/dirty-data-samples-to-practice-on/#8_Invoices_With_Merged_Categories_and_Merged_Amounts).
:::

## Practice - Social Security

::: nonincremental
1.  Navigate to the following link: [here](%22https://www.ssa.gov/oact/babynames/limits.html%22)
2.  Download the state-specific data.
3.  Read in the data from Virginia.
    -   Make some plots for a name of your choosing (time series, distributions, etc.)
4.  Read in *all* of the data
    -   Combine into one data.frame.
    -   What state is your chosen name "most" popular?
    -   What year is your chosen name "most" popular?
    -   Analyze the "gender neutrality" of a relatively unisex name (e.g. "Alex").
    -   Analyze a common name and it's nickname (e.g. "Alex" vs "Alexander").
:::
