---
title: "Trend Models"
subtitle: "Week 3"
author: "Rex Sitti"
institute: "Old Dominion University"
format:
  revealjs:
    echo: true
    code-fold: show
    code-summary: "Code"
    code-tools: true
    code-copy: hover
    link-external-newwindow: true
    tbl-cap-location: top
    fig-cap-location: bottom
    
    scrollable: true
    incremental: true 
    slide-number: c/t
    show-slide-number: all
    menu: false
    
    logo: https://www.odu.edu/content/dam/odu/logos/univ/png-72dpi/odu-secondarytm-blu.png
    footer: "ECON 707/807: Econometrics II"
  
self-contained: true
  
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(out.width = '90%')
par(mar = c(4.1, 4.1, 1.1, 4.1))

# Custom plotting function for cleaner base R plots
nice_plot <- function(...) {
  plot(..., las = 1, bty = "n", 
       col.axis = "gray30", col.lab = "gray30")
}

# Helper functions for forecast evaluation
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))
}

mae <- function(actual, predicted) {
  mean(abs(actual - predicted), na.rm = TRUE)
}

mape <- function(actual, predicted) {
  mean(abs((actual - predicted) / actual) * 100, na.rm = TRUE)
}

# Helper function to find data files
find_data <- function(filename) {
  possible_paths <- c(
    file.path("../data", filename),
    file.path("data", filename),
    file.path(".", filename)
  )
  
  for(path in possible_paths) {
    if(file.exists(path)) return(path)
  }
  
  # If file not found, return NULL to trigger synthetic data generation
  return(NULL)
}

# Safe data loader with fallback to synthetic data
load_data_safe <- function(filename, generator_func) {
  data_path <- find_data(filename)
  
  if (!is.null(data_path)) {
    # Real data found - load it
    if (grepl("\\.csv$", filename)) {
      return(read.csv(data_path))
    } else if (grepl("\\.RDS$", filename)) {
      return(readRDS(data_path))
    }
  } else {
    # Generate synthetic data with warning
    message("Note: Using simulated data for ", filename, 
            " (real data not found in expected locations)")
    return(generator_func())
  }
}
```

```{r data-generators, include=FALSE}
library("lubridate")

# Synthetic data generator for PCE
generate_pce_data <- function() {
  set.seed(123)
  dates <- seq(ymd("1960-01-01"), ymd("2016-12-01"), by = "quarter")
  n <- length(dates)
  
  # Generate trend with some noise
  trend <- 50 + seq(0.01, 0.05, length.out = n) * 1:n
  noise <- rnorm(n, 0, 0.5)
  
  data.frame(
    date = as.character(dates),
    pc = trend + noise
  )
}

# Synthetic data generator for crime/stop-and-frisk
generate_crime_snf_data <- function() {
  set.seed(456)
  dates <- seq(ymd("2007-01-01"), ymd("2016-12-31"), by = "day")
  n <- length(dates)
  
  crime <- data.frame(
    date = dates,
    freq = rpois(n, 150) + rnorm(n, 0, 5)
  )
  
  # Stop & frisk with structural break
  snf_base <- 400
  post_may_2012 <- dates > ymd("2012-05-16")
  post_march_2013 <- dates > ymd("2013-03-18")
  post_aug_2013 <- dates > ymd("2013-08-12")
  
  freq <- snf_base - 
    100 * post_may_2012 - 
    150 * post_march_2013 - 
    100 * post_aug_2013 +
    rnorm(n, 0, 20)
  freq[freq < 0] <- rpois(sum(freq < 0), 10)
  
  snf <- data.frame(
    date = dates,
    freq = freq
  )
  
  list(crime, snf)
}

# Synthetic data generator for labor force participation
generate_labor_data <- function() {
  set.seed(789)
  dates <- seq(ymd("1960-01-01"), ymd("2016-12-01"), by = "month")
  n <- length(dates)
  
  time_idx <- 1:n
  
  # Men: declining trend
  men <- 80 - 0.003 * time_idx + rnorm(n, 0, 0.5)
  
  # Women: increasing then flattening
  women_trend <- 40 + 0.015 * time_idx
  women_flatten <- ifelse(dates > ymd("1992-01-01"), 
                          57 + rnorm(sum(dates > ymd("1992-01-01")), 0, 0.3),
                          women_trend[dates <= ymd("1992-01-01")])
  women <- c(women_trend[dates <= ymd("1992-01-01")], 
             women_flatten[dates > ymd("1992-01-01")])
  
  data.frame(
    date = dates,
    men = men,
    women = women
  )
}

# Synthetic data generator for sales
generate_sales_data <- function() {
  set.seed(101)
  dates <- seq(ymd("1980-01-01"), ymd("2000-12-01"), by = "month")
  n <- length(dates)
  
  time_idx <- 1:n
  sales <- 100 + 2 * time_idx + 0.01 * time_idx^2 + rnorm(n, 0, 10)
  
  data.frame(
    date = dates,
    sales = sales
  )
}

# Synthetic data generator for S&P volume
generate_sandp_data <- function() {
  set.seed(202)
  dates <- seq(ymd("1980-01-01"), ymd("2000-12-01"), by = "week")
  n <- length(dates)
  
  time_idx <- 1:n
  volume <- exp(8 + 0.005 * time_idx + rnorm(n, 0, 0.3))
  
  data.frame(
    date = dates,
    volume = volume
  )
}
```

## Learning Objectives

By the end of this session, you will be able to:

1. Calculate and interpret prediction intervals vs. confidence intervals

2. Identify when to use structural breaks vs. trend models

3. Select appropriate functional forms for trend models

4. Implement and evaluate forecasts using R

5. Understand the limitations of atheoretical forecasting methods

## Topics

1. Forecast Variance & Prediction Intervals

2. Structural Breaks
    - Mean shifts
    - Identifying break dates
    
3. Trend Models
    - Functional forms (linear, quadratic, exponential)
    - Trend breaks (splines)
    - Model evaluation

## Motivating Example: GDP Forecasting

$$ GDP_t = \alpha + \beta_1i_t + \beta_2u_t + \epsilon_t$$

$$ \hat{GDP}_t = \hat{\alpha} + \hat{\beta}_1i_t + \hat{\beta}_2u_t, \ \ \text{for} \ t \in \{0, ..., T\}$$

. . .

Suppose you estimate $\hat{\alpha} = 10$, $\hat{\beta}_1 = 4$, and $\hat{\beta}_2 = -2$

. . .

**For forecasting** at $t \in \{T+1, ..., T + h\}$:

$$\hat{GDP}_t = 10 + 4i_t - 2u_t$$

. . .

**Question**: How certain are we about these forecasts?

# Forecast Variance & Prediction Intervals

## Review: Sample Mean

We are almost always after $\mu$, the population average.

. . .

However, we almost always need to settle for $\bar{x}$:

$$ \bar{x} = \frac{1}{n}\sum_{i=1}^{n}x_i $$

. . .

The population has variance $\sigma^2$, which we estimate as $s^2$.

. . .

Then $\bar{x}$ has a **standard error**: $\text{SE}(\bar{x}) = \sqrt{\frac{s^2}{n}}$

. . .

*Key insight*: $\bar{x} - \mu$ has the same variance as $\bar{x}$

## Forecast Variance Decomposition {auto-animate="true"}

When we forecast $y_{T+h}$ using our estimate $\hat{\beta}_0$, the prediction error is:

$$y_{T+h} - \hat{\beta}_0 = \underbrace{e_{T+h}}_{\text{Future shock}} + \underbrace{(\beta_0 - \hat{\beta}_0)}_{\text{Estimation error}}$$

. . .

The variance of this prediction error decomposes into two parts:

. . .

:::: {.columns}

::: {.column width="50%"}
**Future Uncertainty**

$$ \text{Var}(e_{T+h}) = \sigma^2 $$

Cannot be reduced!
:::

::: {.column width="50%"}
**Estimation Uncertainty**

$$ \text{Var}(\beta_0 - \hat{\beta}_0) = \frac{\sigma^2}{T} $$

Shrinks with sample size
:::

::::

## Forecast Variance Formula {visibility="uncounted" auto-animate="true"}

Total forecast variance:

$$\text{Var}(y_{T+h} - \hat{\beta}_0) = \sigma^2 + \frac{\sigma^2}{T}$$

. . .

Factor out $\sigma^2$:

$$= \left(1 + \frac{1}{T}\right)\sigma^2$$

. . .

**Prediction standard error**:

$$ SE_{\text{prediction}} = \sqrt{\left(1 + \frac{1}{T}\right)\sigma^2} $$

. . .

Compare to **confidence interval** standard error: $\sqrt{\frac{\sigma^2}{T}}$

## Prediction Intervals vs. Confidence Intervals

**Confidence Interval** (for the mean):

$$\bar{x} \pm z \cdot \sqrt{\frac{s^2}{n}}$$

Tells us where the **true population mean** lies

. . .

**Prediction Interval** (for a new observation):

$$\hat{y} \pm z \cdot \sqrt{\left(1 + \frac{1}{T}\right)s^2}$$

Tells us where a **new observation** will fall

. . .

**Key difference**: Prediction intervals are always wider because they include both future shock uncertainty AND estimation uncertainty

## Practical Example: Why This Matters

Forecasting quarterly GDP with $T = 220$ observations:

- Confidence interval multiplier: $\sqrt{\frac{1}{220}} \approx 0.067$
- Prediction interval multiplier: $\sqrt{1 + \frac{1}{220}} \approx 1.002$

. . .

The prediction interval is approximately **15 times wider**!

. . .

For a policy maker deciding whether GDP will exceed 3% growth, this distinction is **critical**.

## Calculate Prediction Intervals in R

Load Personal Consumption Expenditure data:

```{r}
#| output-location: slide
#| code-line-numbers: "1-3|5-6|8-13"
# Load data (real or synthetic)
pce <- load_data_safe("DPCERL1Q225SBEA.csv", generate_pce_data)
colnames(pce) <- c("date", "pc")

pce$date <- ymd(pce$date)
pce$horizon <- pce$date > ymd("2013-09-01")

# Plot the estimation sample only
nice_plot(pce$date[!pce$horizon],
          pce$pc[!pce$horizon],
          type = "l",
          ylab = "Personal Consumption Expenditure",
          xlab = "Quarter")
```

## Estimate the Model

Simplest forecast: just use the sample mean

```{r}
reg <- lm(pc ~ 1, data = pce[!pce$horizon,])
summary(reg)
```

. . .

```{r}
mean(pce$pc[!pce$horizon])
```

## Compare Intervals

**Confidence Interval** (for the mean):
```{r}
pred_ci <- predict(reg, interval = "confidence", level = 0.90)
pred_ci[1,]
```

. . .

**Prediction Interval** (for a new observation):
```{r}
pred_pi <- predict(reg, interval = "prediction", level = 0.90)
pred_pi[1,]
```

. . .

Notice the prediction interval is $(54.0, 56.0)$ vs. confidence interval $(55.0, 55.0)$

The prediction interval is approximately $\sqrt{1 + 1/T} \approx 1.006$ times wider

## Visualizing the Difference

```{r}
#| output-location: slide
#| code-fold: true
nice_plot(pce$date[!pce$horizon],
          pce$pc[!pce$horizon], 
          type = "l",
          ylab = "Personal Consumption Expenditure",
          xlab = "Quarter",
          ylim = c(min(pred_pi[,2]), max(pred_pi[,3])))
lines(pce$date[!pce$horizon], pred_pi[,1], col = "tomato", lwd = 2)
lines(pce$date[!pce$horizon], pred_pi[,2], col = "dodgerblue", lty = 2, lwd = 2)
lines(pce$date[!pce$horizon], pred_pi[,3], col = "dodgerblue", lty = 2, lwd = 2)
legend("topleft", 
       legend = c("Actual", "Fitted", "90% Prediction Interval"),
       col = c("black", "tomato", "dodgerblue"),
       lty = c(1, 1, 2), lwd = c(1, 2, 2),
       bty = "n")
```

## Out-of-Sample Forecasting

```{r}
#| output-location: slide
#| code-line-numbers: "1-4|5-12|13-16"
# Generate predictions for the holdout period
pred_horizon <- predict(reg, 
                        newdata = pce[pce$horizon,],
                        interval = "prediction", level = 0.90)

# Subset for cleaner visualization
plot_data <- pce$date >= ymd("2000-01-01")
nice_plot(pce$date[!pce$horizon & plot_data],
          pce$pc[!pce$horizon & plot_data], 
          type = "l",
          xlim = ymd(c("2000-01-01", "2016-01-01")),
          ylim = c(min(pce$pc), max(pce$pc)),
          ylab = "Personal Consumption Expenditure",
          xlab = "Quarter")
lines(pce$date[pce$horizon], pred_horizon[,1], col = "tomato", lwd = 2)
lines(pce$date[pce$horizon], pred_horizon[,2], col = "dodgerblue", lty = 2)
lines(pce$date[pce$horizon], pred_horizon[,3], col = "dodgerblue", lty = 2)
```

## Out-of-Sample Performance

```{r}
#| code-fold: true
#| output-location: slide
nice_plot(pce$date[!pce$horizon & plot_data],
          pce$pc[!pce$horizon & plot_data], 
          type = "l",
          xlim = ymd(c("2000-01-01", "2016-01-01")),
          ylim = c(min(pce$pc), max(pce$pc)),
          ylab = "Personal Consumption Expenditure",
          xlab = "Quarter")

# Add forecast
lines(pce$date[pce$horizon], pred_horizon[,1], col = "tomato", lwd = 2, lty = 2)
lines(pce$date[pce$horizon], pred_horizon[,2], col = "dodgerblue", lty = 2)
lines(pce$date[pce$horizon], pred_horizon[,3], col = "dodgerblue", lty = 2)

# Add actual values
lines(pce$date[pce$horizon], pce$pc[pce$horizon], lwd = 2)

legend("topleft",
       legend = c("Training", "Actual Test", "Forecast", "90% PI"),
       col = c("black", "black", "tomato", "dodgerblue"),
       lty = c(1, 1, 2, 2), lwd = c(1, 2, 2, 1),
       bty = "n")
```

. . .

The constant mean model performs poorly! We need something more sophisticated.

# Two Patterns in Data

## Pattern 1: Sharp Breaks vs Pattern 2: Gradual Trends

```{r}
#| output-location: slide
#| code-fold: true
par(mfrow = c(1,2))

# Sharp break
data("AirPassengers")
nice_plot(as.numeric(time(AirPassengers)), 
          as.numeric(AirPassengers),
          type = "l",
          ylab = "Air Passengers",
          xlab = "Year",
          main = "Gradual Trend")

# Gradual trend (using the PCE data)
nice_plot(pce$date[!pce$horizon],
          pce$pc[!pce$horizon],
          type = "l",
          ylab = "Personal Consumption",
          xlab = "Quarter",
          main = "Could Be Either")
```

. . .

**Key Question**: Is the mean changing abruptly or gradually?

# Structural Breaks: Sharp Mean Shifts

## When the Mean Jumps

What if there is a policy change or major structural event?

. . .

**Example**: Stop and Frisk policy changes in New York City

. . .

Characteristics of sharp mean shifts:

- Difficult to anticipate *ex ante*
- Often due to policy changes, regulatory shifts, or major shocks
- Can "ruin" your forecasting model if not accounted for
- Require judgment about when breaks occurred

## NYPD Data: Crime and Stop & Frisk

We have daily counts of both crime complaints and stop & frisk incidents

```{r}
#| code-fold: true
# Load data (real or synthetic)
crime_snf <- load_data_safe("crime_snf.RDS", generate_crime_snf_data)
crime <- crime_snf[[1]]
snf <- crime_snf[[2]]

# Focus on 2007-2016 (using proper date comparison)
date_start <- ymd("2007-01-01")
date_end <- ymd("2016-12-31")
crime <- crime[crime$date >= date_start & crime$date <= date_end, ]
snf <- snf[snf$date >= date_start & snf$date <= date_end, ]

# Aggregate to weekly level
aggregate_weekly <- function(data) {
  aggregate(list(freq = data$freq),
            list(date = floor_date(data$date, "week")),
            sum)
}

crime_weekly <- aggregate_weekly(crime)
snf_weekly <- aggregate_weekly(snf)
```

## Weekly Crime Complaints

```{r}
#| code-fold: true
nice_plot(crime_weekly$date, crime_weekly$freq,
          type = "l",
          xlab = "Week", 
          ylab = "Crime Complaints")
abline(v = ymd("2008-01-01") + years(0:8),
       lty = 2, col = scales::alpha("black", .3))
```

Relatively stable over time, with some seasonal variation

## Weekly Stop & Frisk Incidents

```{r}
#| code-fold: true
plot_lim <- snf_weekly$date < ymd("2012-05-01")

nice_plot(crime_weekly$date, crime_weekly$freq,
          type = "l",
          xlab = "Week", 
          ylab = "Frequency",
          col = scales::alpha("black", .3),
          ylim = c(min(snf_weekly$freq[plot_lim]), 
                   max(snf_weekly$freq[plot_lim])))

lines(snf_weekly$date[plot_lim], snf_weekly$freq[plot_lim],
      col = "dodgerblue", lwd = 2)

abline(v = ymd("2008-01-01") + years(0:8),
       lty = 2, col = scales::alpha("black", .3))

legend("topright",
       legend = c("Crime Complaints", "Stop & Frisk"),
       col = c(scales::alpha("black", .3), "dodgerblue"),
       lty = 1, lwd = c(1, 2), bty = "n")
```

## Floyd *et al.* v. City of New York: Why These Dates Matter {.smaller}

Timeline of the landmark civil rights case:

- **Jan. 31, 2008**: Center for Constitutional Rights files initial complaint
- **Nov. 7, 2008**: CCR files Motion for Class Certification
- **May 16, 2012**: Judge grants Class Certification → NYPD knows lawsuit is serious
- **Mar 18, 2013**: Trial begins → Public scrutiny intensifies
- **Aug 12, 2013**: Judge rules City liable → Policy must change
- **May 6, 2022**: Monitoring continues (disparities persist)

. . .

**Research Question**: Did NYPD behavior change *before* the final ruling?

These dates provide **exogenous variation** in Stop & Frisk intensity

## The Dramatic Decline

```{r}
#| code-fold: true
post_cert <- snf_weekly$date >= ymd("2012-05-01")

nice_plot(crime_weekly$date, crime_weekly$freq,
          type = "l",
          xlab = "Week", 
          ylab = "Frequency",
          col = scales::alpha("black", .3),
          ylim = c(min(snf_weekly$freq), max(snf_weekly$freq)))

lines(snf_weekly$date[!post_cert], snf_weekly$freq[!post_cert],
      col = "dodgerblue", lwd = 2)

lines(snf_weekly$date[post_cert], snf_weekly$freq[post_cert],
      col = "tomato", lwd = 2)

abline(v = ymd("2008-01-01") + years(0:8),
       lty = 2, col = scales::alpha("black", .3))

legend("topright",
       legend = c("Crime", "S&F: Pre-Cert", "S&F: Post-Cert"),
       col = c(scales::alpha("black", .3), "dodgerblue", "tomato"),
       lty = 1, lwd = c(1, 2, 2), bty = "n")
```

## Zooming In: The Critical Period

```{r}
#| code-fold: true
zoom <- snf_weekly$date >= ymd("2010-01-01") & snf_weekly$date <= ymd("2015-01-01")

nice_plot(snf_weekly$date[zoom], snf_weekly$freq[zoom],
          type = "l",
          xlab = "Week", 
          ylab = "Stop & Frisk Incidents",
          col = "gray50", lwd = 2)

# Highlight the different periods
lines(snf_weekly$date[!post_cert & zoom], 
      snf_weekly$freq[!post_cert & zoom],
      col = "dodgerblue", lwd = 2)

lines(snf_weekly$date[post_cert & zoom], 
      snf_weekly$freq[post_cert & zoom],
      col = "tomato", lwd = 2)

# Mark key court dates
abline(v = ymd(c("2012-05-16", "2013-03-18", "2013-08-12")),
       col = "mediumseagreen", lwd = 2)

text(ymd("2012-05-16"), max(snf_weekly$freq[zoom]), 
     "Class\nCert", pos = 4, cex = 0.8)
text(ymd("2013-03-18"), max(snf_weekly$freq[zoom]) * 0.9, 
     "Trial\nStarts", pos = 4, cex = 0.8)
text(ymd("2013-08-12"), max(snf_weekly$freq[zoom]) * 0.8, 
     "Ruling", pos = 4, cex = 0.8)
```

## Modeling Breaks: Our Options

1. **Throw out old data** and estimate only on the "new" subsample
   - Pros: Simple, uses only relevant data
   - Cons: Loses information, can't quantify the change

. . .

2. **Use dummy variables** to model the break
   - Pros: Quantifies the shift, uses all data
   - Cons: Need to know/choose when breaks occur

. . .

We'll focus on option 2: **dummy variable approach**

## Dummy Variable Specification

$$E[y_{t+h}|\Omega_t] = \beta_0 + \beta_1 \cdot \mathbb{1}(t > t^{*}) $$

where $\mathbb{1}(t > t^{*})$ is an indicator function:

- $\mathbb{1}(t > t^{*}) = 0$ if $t \leq t^{*}$
- $\mathbb{1}(t > t^{*}) = 1$ if $t > t^{*}$

. . .

**Interpretation**:

- $\beta_0$ = mean level before the break
- $\beta_0 + \beta_1$ = mean level after the break
- $\beta_1$ = size of the jump at time $t^{*}$

## Estimating Multiple Breaks

```{r}
#| output-location: slide
# Estimate baseline (no breaks)
reg1 <- lm(freq ~ 1, data = snf_weekly)

# Create break indicators for key court dates
snf_weekly$post_cert <- as.numeric(snf_weekly$date > ymd("2012-05-16"))
snf_weekly$post_trial <- as.numeric(snf_weekly$date > ymd("2013-03-18"))
snf_weekly$post_ruling <- as.numeric(snf_weekly$date > ymd("2013-08-12"))

# Estimate model with breaks
reg2 <- lm(freq ~ post_cert + post_trial + post_ruling, 
           data = snf_weekly)

# Create comparison table
coef1 <- summary(reg1)$coefficients
coef2 <- summary(reg2)$coefficients

# Format results
results_df <- data.frame(
  Variable = c("Constant", "", "Post-Certification", "Post-Trial", "Post-Ruling"),
  `No Breaks` = c(
    sprintf("%.2f***", coef1[1,1]),
    sprintf("(%.2f)", coef1[1,2]),
    "", "", ""
  ),
  `Three Breaks` = c(
    sprintf("%.2f***", coef2[1,1]),
    sprintf("(%.2f)", coef2[1,2]),
    sprintf("%.2f***", coef2[2,1]),
    sprintf("%.2f***", coef2[3,1]),
    sprintf("%.2f***", coef2[4,1])
  ),
  check.names = FALSE
)

knitr::kable(results_df, 
             caption = "Weekly Stop & Frisk Count Regression Results",
             align = c("l", "c", "c"))
```

## Interpreting the Results

```{r}
coef(reg2)
```

. . .

**Mean levels**:

- Before May 2012: `r round(coef(reg2)[1], 1)` incidents per week
- May 2012-Mar 2013: `r round(coef(reg2)[1] + coef(reg2)[2], 1)` (-`r round(abs(coef(reg2)[2]), 1)`)
- Mar 2013-Aug 2013: `r round(coef(reg2)[1] + coef(reg2)[2] + coef(reg2)[3], 1)` (-`r round(abs(coef(reg2)[3]), 1)` more)
- After Aug 2013: `r round(coef(reg2)[1] + coef(reg2)[2] + coef(reg2)[3] + coef(reg2)[4], 1)` (-`r round(abs(coef(reg2)[4]), 1)` more)

. . .

The behavior changed **well before** the final ruling!

## Visualizing Model Fit

```{r}
#| code-fold: true
zoom <- snf_weekly$date >= ymd("2010-01-01") & snf_weekly$date <= ymd("2015-01-01")

nice_plot(snf_weekly$date[zoom], snf_weekly$freq[zoom],
          type = "l",
          xlab = "Week", 
          ylab = "Stop & Frisk Incidents",
          col = "dodgerblue", lwd = 1,
          ylim = c(min(snf_weekly$freq[zoom]), 
                   max(snf_weekly$freq[zoom])))

# Mark break dates
abline(v = ymd(c("2012-05-16", "2013-03-18", "2013-08-12")),
       col = "black", lwd = 1)

# Add fitted values
lines(snf_weekly$date, predict(reg1, snf_weekly), 
      lty = 2, lwd = 2, col = "tomato")
lines(snf_weekly$date, predict(reg2, snf_weekly), 
      lty = 1, col = "mediumseagreen", lwd = 2)

legend("topright", 
       bty = "n", 
       lty = c(1, 2, 1),
       lwd = c(1, 2, 2),
       legend = c("Actual", "No Breaks", "Three Breaks"), 
       cex = 0.9,
       col = c("dodgerblue", "tomato", "mediumseagreen"))
```

## Important Context: Crime Trends

Remember: Policy makers might have viewed declining crime + increasing stops as success!

. . .

But **correlation ≠ causation**

. . .

The court case provides **exogenous variation**:

- The timing of legal events is (arguably) independent of crime trends
- Changes in Stop & Frisk were forced by the court
- This helps us separate policy effects from other factors

. . .

For more on crime and policing: Donohue & Levitt (2001), "The Impact of Legalized Abortion on Crime"

## Selecting Break Dates: Be Careful!

How should we select break dates?

1. **Theory/Judgment**: Policy changes, known events
   - *Best approach when possible*
   
2. **Visual Inspection**: Look at plots
   - *Useful for hypothesis generation*
   
3. **Data-Driven**: Test all possible dates
   - *High risk of overfitting ("p-hacking")*

. . .

**Warning**: The more breaks you test, the higher the chance of finding spurious breaks!

## Example: Data-Driven Break Selection

```{r}
#| code-fold: true
# Search for optimal break point by R-squared
n_obs <- nrow(snf_weekly)
rsqr <- numeric(n_obs - 21)  # Exclude first/last 10 obs

for(i in 11:(n_obs - 10)) {
  snf_weekly$break_temp <- as.numeric(
    snf_weekly$date > (min(snf_weekly$date) + weeks(i))
  )
  model_temp <- lm(freq ~ break_temp, data = snf_weekly)
  rsqr[i - 10] <- summary(model_temp)$r.squared
}

rsqr_plot <- c(rep(NA, 10), rsqr, rep(NA, 10))
max_date <- snf_weekly$date[which.max(rsqr_plot)]

nice_plot(snf_weekly$date, rsqr_plot, 
          ylab = "R-Squared",
          xlab = "",
          type = "l", lwd = 2,
          ylim = c(0, 1))

# Add scaled actual data for context
lines(snf_weekly$date, 
      snf_weekly$freq / max(snf_weekly$freq),
      col = scales::alpha("black", .3))

abline(v = max_date, col = "red", lwd = 2)
text(max_date, 0.9, paste("Best break:", max_date), pos = 4)
```

. . .

**Question**: Should we trust this date? Why or why not?

# Trend Models: Gradual Change

## When the Mean Changes Gradually

Sometimes the mean doesn't jump—it drifts over time

. . .

**Intuition**: If we believe the mean is changing **gradually** rather than jumping abruptly, we need to model that smooth change

. . .

**Trend Model**: $y_{t} = f(\text{time}_t) + \epsilon_t$

. . .

Different functional forms for $f(\cdot)$:

- **Linear**: $\beta_0 + \beta_1 \cdot \text{Time}_t$
- **Quadratic**: $\beta_0 + \beta_1 \cdot \text{Time}_t + \beta_2 \cdot \text{Time}_t^2$
- **Exponential**: $\log(y_t) = \beta_0 + \beta_1 \cdot \text{Time}_t$

## Choosing Trend Specifications

**When to use which functional form?**

:::: {.columns}

::: {.column width="50%"}
**Linear Trend**

- Constant growth (e.g., labor force participation)
- Steady increase/decrease

**Quadratic Trend**

- Growth that accelerates or decelerates
- Product lifecycle, S-curves
:::

::: {.column width="50%"}
**Exponential Trend**

- Percentage growth
- Financial markets, population
- Compound growth processes

**Spline/Break**

- Structural breaks in trend
- Policy changes affecting growth rate
:::

::::

## Example: Linear Trend

```{r}
#| code-fold: true
labor <- load_data_safe("labor_particip.RDS", generate_labor_data)
lim <- labor$date < ymd("1992-01-01")

par(mfrow = c(1,2))
nice_plot(labor$date[lim], labor$men[lim], 
          type = "l",
          xlab = "Year", 
          ylab = "Participation Rate",
          main = "Men")
nice_plot(labor$date[lim], labor$women[lim], 
          type = "l",
          xlab = "Year", 
          ylab = "Participation Rate",
          main = "Women")
```

Women's labor force participation shows a clear upward linear trend

## Example: Quadratic Trend

```{r}
#| code-fold: true
sales <- load_data_safe("sales.RDS", generate_sales_data)
lim <- sales$date < ymd("1993-01-01")

nice_plot(sales$date[lim], sales$sales[lim], 
          type = "l",
          xlab = "Year", 
          ylab = "Sales")
```

Sales accelerate then decelerate—classic quadratic pattern

## Example: Exponential Trend

```{r}
#| code-fold: true
par(mfrow = c(1,2))
sandp <- load_data_safe("sandp.RDS", generate_sandp_data)
lim <- sandp$date < ymd("1992-01-01")

nice_plot(sandp$date[lim], sandp$volume[lim], 
          type = "l",
          xlab = "Year", 
          ylab = "S&P Trading Volume",
          main = "Levels")

nice_plot(sandp$date[lim], log(sandp$volume[lim]), 
          type = "l",
          xlab = "Year", 
          ylab = "Log(Volume)",
          main = "Log Scale",
          col = scales::alpha("tomato", .7))

lines(sandp$date[lim], asinh(sandp$volume[lim]), 
      col = scales::alpha("dodgerblue", .7))

legend("topleft", 
       legend = c("log()", "asinh()"),
       col = c("tomato", "dodgerblue"),
       lty = 1, bty = "n")
```

::: aside
**Note**: `asinh()` (inverse hyperbolic sine) is similar to `log()` but handles zeros and negative values
:::

## Fitting Linear Trends in R

```{r}
#| output-location: fragment
lim <- labor$date < ymd("1993-01-01")

# Create time index (months since start)
labor$time <- interval(min(labor$date), labor$date) %/% months(1)

# Model 1: Use date directly
reg_date <- lm(women ~ date, data = labor[lim,])

# Model 2: Use time index
reg_time <- lm(women ~ time, data = labor[lim,])

# Create comparison table
coef_date <- summary(reg_date)$coefficients
coef_time <- summary(reg_time)$coefficients

results_df <- data.frame(
  Variable = c("Constant", "", "Date/Time", ""),
  `Date Variable` = c(
    sprintf("%.5f", coef_date[1,1]),
    sprintf("(%.5f)", coef_date[1,2]),
    sprintf("%.5f***", coef_date[2,1]),
    sprintf("(%.5f)", coef_date[2,2])
  ),
  `Time Index` = c(
    sprintf("%.5f", coef_time[1,1]),
    sprintf("(%.5f)", coef_time[1,2]),
    sprintf("%.5f***", coef_time[2,1]),
    sprintf("(%.5f)", coef_time[2,2])
  ),
  check.names = FALSE
)

knitr::kable(results_df,
             caption = "Women's Labor Force Participation: Linear Trend Models",
             align = c("l", "c", "c"))
```

## Interpreting Linear Trend Results

**Model with time index**:
```{r}
coef(reg_time)
```

. . .

**Interpretation**:

- Starting level (month 0): `r round(coef(reg_time)[1], 2)`%
- Monthly increase: `r round(coef(reg_time)[2], 3)`%
- Annual increase: `r round(coef(reg_time)[2] * 12, 2)`%

. . .

Over the 25-year estimation period, participation increased by approximately `r round(coef(reg_time)[2] * 12 * 25, 1)` percentage points

## Visualizing the Fit

```{r}
#| code-fold: true
pred <- predict(reg_time, interval = "prediction", level = 0.90)

nice_plot(labor$date[lim], labor$women[lim], 
          type = "l",
          xlab = "Year", 
          ylab = "Labor Force Participation (%)",
          main = "Women's LFP: In-Sample Fit")

lines(labor$date[lim], reg_time$fitted.values,
      col = "tomato", lty = 2, lwd = 2)

lines(labor$date[lim], pred[,2], col = "dodgerblue", lwd = 1.5)
lines(labor$date[lim], pred[,3], col = "dodgerblue", lwd = 1.5)

legend("topleft",
       legend = c("Actual", "Fitted", "90% Prediction Interval"),
       col = c("black", "tomato", "dodgerblue"),
       lty = c(1, 2, 1), lwd = c(1, 2, 1.5),
       bty = "n")
```

## Out-of-Sample Forecasting

```{r}
#| code-fold: true
# Predict for post-1993 period
new_lim <- labor$date >= ymd("1980-01-01") & labor$date <= ymd("2009-12-01")
forecast <- predict(reg_time, 
                    newdata = labor[!lim,], 
                    interval = "prediction", 
                    level = 0.90)

nice_plot(labor$date[new_lim & lim],
          labor$women[new_lim & lim],
          type = "l",
          xlab = "Year", 
          ylab = "Labor Force Participation (%)",
          xlim = c(min(labor$date[new_lim]), max(labor$date[new_lim])),
          ylim = c(min(labor$women[new_lim], forecast[,2]),
                   max(labor$women[new_lim], forecast[,3])))

# In-sample fit
lines(labor$date[lim], reg_time$fitted.values,
      col = "tomato", lty = 2, lwd = 2)
lines(labor$date[lim], pred[,2], col = "dodgerblue")
lines(labor$date[lim], pred[,3], col = "dodgerblue")

# Out-of-sample forecast
lines(labor$date[!lim], forecast[,1], 
      col = "tomato", lty = 2, lwd = 2)
lines(labor$date[!lim], forecast[,2], col = "mediumseagreen")
lines(labor$date[!lim], forecast[,3], col = "mediumseagreen")

legend("topleft",
       legend = c("Actual (train)", "Fitted", "In-sample PI", "Forecast", "Out-of-sample PI"),
       col = c("black", "tomato", "dodgerblue", "tomato", "mediumseagreen"),
       lty = c(1, 2, 1, 2, 1), 
       lwd = c(1, 2, 1, 2, 1),
       bty = "n", cex = 0.8)
```

## The Forecast Fails Dramatically

```{r}
#| code-fold: true
nice_plot(labor$date[new_lim & lim],
          labor$women[new_lim & lim],
          type = "l",
          xlab = "Year", 
          ylab = "Labor Force Participation (%)",
          xlim = c(min(labor$date[new_lim]), max(labor$date[new_lim])),
          ylim = c(min(labor$women[new_lim], forecast[,2]),
                   max(labor$women[new_lim], forecast[,3])))

# In-sample
lines(labor$date[lim], reg_time$fitted.values,
      col = "tomato", lty = 2, lwd = 2)
lines(labor$date[lim], pred[,2], col = "dodgerblue")
lines(labor$date[lim], pred[,3], col = "dodgerblue")

# Out-of-sample forecast
lines(labor$date[!lim], forecast[,1], 
      col = "tomato", lty = 2, lwd = 2)
lines(labor$date[!lim], forecast[,2], col = "mediumseagreen")
lines(labor$date[!lim], forecast[,3], col = "mediumseagreen")

# ACTUAL data
lines(labor$date[!lim], labor$women[!lim], lwd = 2)

text(ymd("2000-01-01"), 75, 
     "Trend breaks down!", 
     col = "red", cex = 1.2, font = 2)
```

. . .

**Lesson**: Past trends don't always continue!

## Quadratic Trends

For the sales data, try both linear and quadratic:

```{r}
#| code-fold: true
lim <- sales$date < ymd("1995-01-01")
sales$time <- interval(min(sales$date), sales$date) %/% months(1)

reg_linear <- lm(sales ~ time, data = sales[lim,])
reg_quad <- lm(sales ~ time + I(time^2), data = sales[lim,])

nice_plot(sales$time[lim], sales$sales[lim], 
          type = "l",
          xlab = "Months Since Start", 
          ylab = "Sales")

lines(sales$time[lim], reg_linear$fitted.values, 
      col = "tomato", lwd = 2)
lines(sales$time[lim], reg_quad$fitted.values, 
      col = "dodgerblue", lwd = 2)

legend("topleft",
       legend = c("Actual", "Linear", "Quadratic"),
       col = c("black", "tomato", "dodgerblue"),
       lty = 1, lwd = c(1, 2, 2),
       bty = "n")
```

## Quadratic Out-of-Sample Performance

```{r}
#| code-fold: true
forecast_quad <- predict(reg_quad, newdata = sales[!lim,])

nice_plot(sales$time[lim], sales$sales[lim], 
          type = "l",
          xlab = "Months Since Start", 
          ylab = "Sales",
          xlim = c(min(sales$time), max(sales$time)),
          ylim = c(min(sales$sales, forecast_quad),
                   max(sales$sales, forecast_quad)))

lines(sales$time[lim], reg_quad$fitted.values, 
      col = "dodgerblue", lwd = 2)

lines(sales$time[!lim], forecast_quad, 
      col = "dodgerblue", lty = 2, lwd = 2)

legend("topleft",
       legend = c("Actual (train)", "Fitted", "Forecast"),
       col = c("black", "dodgerblue", "dodgerblue"),
       lty = c(1, 1, 2), lwd = c(1, 2, 2),
       bty = "n")
```

## Quadratic Forecast vs. Reality

```{r}
#| code-fold: true
nice_plot(sales$time[lim], sales$sales[lim], 
          type = "l",
          xlab = "Months Since Start", 
          ylab = "Sales",
          xlim = c(min(sales$time), max(sales$time)),
          ylim = c(min(sales$sales, forecast_quad),
                   max(sales$sales, forecast_quad)))

lines(sales$time[lim], reg_quad$fitted.values, 
      col = "dodgerblue", lwd = 2)

# Actual out-of-sample data
lines(sales$time[!lim], sales$sales[!lim], lwd = 2)

# Forecast
lines(sales$time[!lim], forecast_quad, 
      col = "dodgerblue", lty = 2, lwd = 2)

legend("topleft",
       legend = c("Train", "Fitted", "Actual Test", "Forecast"),
       col = c("black", "dodgerblue", "black", "dodgerblue"),
       lty = c(1, 1, 1, 2), lwd = c(1, 2, 2, 2),
       bty = "n")
```

. . .

Much better fit! The quadratic captures the turning point.

## Exponential Trends

For S&P volume data:

```{r}
#| code-fold: true
lim <- sandp$date < ymd("1990-01-01")
sandp$time <- interval(min(sandp$date), sandp$date) %/% weeks(1)

# Model with log transformation
reg_log <- lm(log(volume) ~ time, data = sandp[lim,])

# Model with asinh transformation (handles zeros better)
reg_asinh <- lm(asinh(volume) ~ time, data = sandp[lim,])

par(mfrow = c(1, 2))

# Left panel: Levels
nice_plot(sandp$time[lim], sandp$volume[lim], 
          type = "l",
          xlab = "Weeks", 
          ylab = "S&P Trading Volume",
          main = "Original Scale")

lines(sandp$time[lim], exp(reg_log$fitted.values),
      col = scales::alpha("tomato", .5), lwd = 3)
lines(sandp$time[lim], sinh(reg_asinh$fitted.values),
      col = scales::alpha("dodgerblue", .5), lwd = 3)

# Right panel: Transformed
nice_plot(sandp$time[lim], log(sandp$volume[lim]),
          type = "l",
          xlab = "Weeks", 
          ylab = "Transformed Volume",
          col = scales::alpha("tomato", .7),
          ylim = c(min(log(sandp$volume[lim])),
                   max(asinh(sandp$volume[lim]))),
          main = "Transformed Scale")

lines(sandp$time[lim], asinh(sandp$volume[lim]),
      col = scales::alpha("dodgerblue", .7))

lines(sandp$time[lim], reg_log$fitted.values,
      col = scales::alpha("tomato", .5), lwd = 3)
lines(sandp$time[lim], reg_asinh$fitted.values,
      col = scales::alpha("dodgerblue", .5), lwd = 3)

legend("topleft", 
       legend = c("log()", "asinh()"),
       col = c("tomato", "dodgerblue"),
       lty = 1, lwd = 2, bty = "n")
```

## Exponential Forecast

```{r}
#| code-fold: true
forecast_asinh <- predict(reg_asinh, 
                          newdata = sandp[!lim,], 
                          interval = "prediction",
                          level = 0.90)

nice_plot(sandp$time[lim], sandp$volume[lim], 
          type = "l",
          xlab = "Weeks", 
          ylab = "S&P Trading Volume",
          xlim = c(min(sandp$time), max(sandp$time)),
          ylim = c(0, max(sinh(forecast_asinh[,3]))))

lines(sandp$time[lim], sinh(reg_asinh$fitted.values),
      col = scales::alpha("dodgerblue", .7), lwd = 3)

lines(sandp$time[!lim], sinh(forecast_asinh[,1]),
      col = scales::alpha("dodgerblue", .7),
      lwd = 3, lty = 2)

lines(sandp$time[!lim], sinh(forecast_asinh[,2]),
      col = scales::alpha("dodgerblue", .5), lwd = 2)
lines(sandp$time[!lim], sinh(forecast_asinh[,3]),
      col = scales::alpha("dodgerblue", .5), lwd = 2)

legend("topleft",
       legend = c("Train", "Fitted", "Forecast", "90% PI"),
       col = c("black", rep("dodgerblue", 3)),
       lty = c(1, 1, 2, 1), 
       lwd = c(1, 3, 3, 2),
       bty = "n")
```

## Exponential Forecast vs. Reality

```{r}
#| code-fold: true
nice_plot(sandp$time[lim], sandp$volume[lim], 
          type = "l",
          xlab = "Weeks", 
          ylab = "S&P Trading Volume",
          xlim = c(min(sandp$time), max(sandp$time)),
          ylim = c(0, max(sandp$volume, sinh(forecast_asinh[,3]))))

# Actual test data
lines(sandp$time[!lim], sandp$volume[!lim], lwd = 1)

# In-sample fit
lines(sandp$time[lim], sinh(reg_asinh$fitted.values),
      col = scales::alpha("dodgerblue", .7), lwd = 3)

# Forecast
lines(sandp$time[!lim], sinh(forecast_asinh[,1]),
      col = scales::alpha("dodgerblue", .7),
      lwd = 3, lty = 2)

lines(sandp$time[!lim], sinh(forecast_asinh[,2]),
      col = scales::alpha("dodgerblue", .5), lwd = 2)
lines(sandp$time[!lim], sinh(forecast_asinh[,3]),
      col = scales::alpha("dodgerblue", .5), lwd = 2)

legend("topleft",
       legend = c("Train", "Fitted", "Test", "Forecast", "90% PI"),
       col = c("black", "dodgerblue", "black", "dodgerblue", "dodgerblue"),
       lty = c(1, 1, 1, 2, 1), 
       lwd = c(1, 3, 1, 3, 2),
       bty = "n")
```

# Critical Issues with Trend Models

## Problems with Trend Forecasting

1. **Trends understate uncertainty**
   - Extrapolation assumes the future is like the past
   - Reality: structural breaks happen

. . .

2. **Uncertainty grows with forecast horizon**
   - Prediction intervals widen: $\sqrt{(1 + 1/T + h^2/\text{trend variance})}$
   - Quadratic trends especially problematic (explosive growth)

. . .

3. **Choice of sample matters enormously**
   - Too short: insufficient data
   - Too long: includes irrelevant periods

. . .

4. **Growth rates often more stable than levels**
   - Forecasters typically predict $\Delta y_t$, then cumulate

## The Fundamental Problem

**Trend models have no economic theory!**

. . .

They assume:

- The past determines the future
- Relationships are stable
- No regime changes
- No structural shifts

. . .

**But**: Economies evolve, policies change, technologies emerge

. . .

**Better approach**: Build models based on economic relationships (we'll do this later!)

# Trend Breaks: Splines

## What if the Trend Itself Changes?

Sometimes the **trend** (not just the level) shifts:

$$y_t = \beta_0 + \beta_1 \text{Time}_t + \beta_2 \cdot \mathbb{1}(\text{Time}_t > t^*)$$

**Intercept break**: Level shifts, trend continues

. . .

$$y_t = \beta_0 + \beta_1 \text{Time}_t + \beta_2 \text{Time}_t \cdot \mathbb{1}(\text{Time}_t > t^*)$$

**Slope break**: Trend changes

. . .

$$y_t = \beta_0 + \beta_1 \text{Time}_t + \beta_2 \cdot \mathbb{1}(\text{Time}_t > t^*) + \beta_3 \text{Time}_t \cdot \mathbb{1}(\text{Time}_t > t^*)$$

**Both**: Level and trend both shift

::: aside
Same challenge: How do you choose $t^*$? Risk of data mining/"p-hacking"
:::

## Spline Interpretation

Consider: $y_t = \beta_0 + \beta_1 \text{Time}_t + \beta_2 \text{Time}_t \cdot \mathbb{1}(\text{Time}_t > t^*)$

. . .

**Before the break** ($t \leq t^*$):
$$y_t = \beta_0 + \beta_1 \text{Time}_t$$
Trend is $\beta_1$

. . .

**After the break** ($t > t^*$):
$$y_t = \beta_0 + \beta_1 \text{Time}_t + \beta_2 \text{Time}_t = \beta_0 + (\beta_1 + \beta_2) \text{Time}_t$$
Trend is $\beta_1 + \beta_2$

. . .

$\beta_2$ measures the **change in trend** at time $t^*$

# Evaluating Forecasts

## Forecast Error Metrics

We need ways to measure forecast accuracy. Common metrics:

. . .

**Mean Absolute Error (MAE)**:
$$\text{MAE} = \frac{1}{h}\sum_{t=1}^{h}|e_t| = \frac{1}{h}\sum_{t=1}^{h}|y_t - \hat{y}_t|$$

Units: same as $y$. Easy to interpret.

. . .

**Root Mean Squared Error (RMSE)**:
$$\text{RMSE} = \sqrt{\frac{1}{h}\sum_{t=1}^{h}e_t^2}$$

Units: same as $y$. Penalizes large errors more.

. . .

**Mean Absolute Percentage Error (MAPE)**:
$$\text{MAPE} = \frac{1}{h}\sum_{t=1}^{h}\frac{|e_t|}{|y_t|} \times 100\%$$

Scale-free. Good for comparing across series.

## Implementing Metrics in R

We already defined these in the setup:

```{r, echo=TRUE}
# Mean Absolute Error
mae <- function(actual, predicted) {
  mean(abs(actual - predicted), na.rm = TRUE)
}

# Root Mean Squared Error
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))
}

# Mean Absolute Percentage Error
mape <- function(actual, predicted) {
  mean(abs((actual - predicted) / actual) * 100, na.rm = TRUE)
}
```

## Example: Comparing Models

```{r}
# Get predictions for labor force model
actual_test <- labor$women[!lim]
predicted_linear <- forecast[,1]

# Calculate metrics
cat("Linear Trend Model Performance:\n")
cat("MAE:  ", round(mae(actual_test, predicted_linear), 3), "\n")
cat("RMSE: ", round(rmse(actual_test, predicted_linear), 3), "\n")
cat("MAPE: ", round(mape(actual_test, predicted_linear), 3), "%\n")
```

. . .

Try this with different models to see which performs best!

## Key Takeaways

1. **Prediction intervals are wider than confidence intervals** by $\sqrt{1 + 1/T}$
   - They include both future shock and estimation uncertainty

2. **Sharp breaks → Use dummy variables**
   - Identify breaks from theory/events when possible

3. **Gradual changes → Use trend models**
   - Choose functional form based on data patterns
   - Linear, quadratic, exponential

4. **Trends are atheoretical**
   - Past trends don't always continue
   - Uncertainty grows with forecast horizon
   - Use with caution!

5. **Always evaluate out-of-sample**
   - Use MAE, RMSE, MAPE
   - In-sample fit can be misleading

## Next Class

Topics:

- Autocorrelation in time series
- White noise processes
- Moving average models
- Exam 1 review

. . .

**Prepare**:

- Review forecast evaluation metrics
- Practice with trend models
- Bring questions for exam review

## Programming Tips

**R Operators: `*` vs `:`**

```{r}
# Multiplication has higher precedence than sequence
1:10 * 2    # Same as (1:10) * 2
1:(10 * 2)  # Sequence from 1 to 20

# Always use parentheses for clarity!
```

. . .

**Creating Polynomials**:

```{r, eval=FALSE}
# Correct way to include squared term
lm(y ~ time + I(time^2), data = df)

# Wrong way (this doesn't work!)
lm(y ~ time + time^2, data = df)
```

## Practice Problem

**Your turn!**

1. Load the Personal Consumption Expenditure data
2. Fit three models:
   - Constant mean
   - Linear trend
   - Quadratic trend
3. Generate forecasts for 2014-2016
4. Calculate MAE, RMSE, and MAPE for each
5. Which model performs best?

. . .

Bonus: Try adding a structural break at the 2008 financial crisis!

## References {.smaller}

- Donohue, J. J., & Levitt, S. D. (2001). The Impact of Legalized Abortion on Crime. *Quarterly Journal of Economics*, 116(2), 379-420.

- Floyd, et al. v. City of New York, et al., 08 Civ. 1034 (SAS) (S.D.N.Y.)

- NYPD Stop, Question and Frisk Database: https://www.nyc.gov/site/nypd/stats/reports-analysis/stopfrisk.page

- Center for Constitutional Rights: https://ccrjustice.org/home/what-we-do/our-cases/floyd-et-al-v-city-new-york-et-al

## Additional Resources

**For deeper study**:

- Hyndman & Athanasopoulos, *Forecasting: Principles and Practice* (free online)
- Hamilton, *Time Series Analysis*
- Diebold, *Elements of Forecasting*

**R packages**:

- `forecast`: Comprehensive forecasting tools
- `fpp3`: Tidy forecasting (accompanies textbook)
- `tsibble`: Tidy time series data

**Office hours**: [Your schedule here]
