---
title: "Vector Autoregression"
subtitle: "Week 12"
author: "Rex Sitti"
institute: "Old Dominion University"
format:
  revealjs:
    echo: true
    code-fold: show
    code-summary: "Code"
    code-tools: true
    code-copy: hover
    link-external-newwindow: true
    tbl-cap-location: top
    fig-cap-location: bottom
    
    scrollable: true
    incremental: true 
    slide-number: c/t
    show-slide-number: all
    menu: false
    
    logo: "https://ww1.odu.edu/facultystaff/communication/toolkit/current-logos/_jcr_content/par/section_1593932067/columns_527553970/column_2/image.img.440.jpg/1599596551953.jpg"
    footer: "ECON 707/807: Econometrics II"
    
self-contained: true

editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(out.width = '90%')

par(mar = c(4.1, 4.1, 1.1, 4.1))
set.seed(321)
library("stargazer")
library("lubridate")
library("forecast")
library("scales")
library("tseries")
library("vars")

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, x)
  } else x
}

# ============================================================================
# SYNTHETIC DATA GENERATORS - Slides work without external data files
# ============================================================================

generate_la_air_quality <- function() {
  # Generate weekly LA air quality data (1970-1979, 508 weeks)
  set.seed(456)
  n <- 508
  
  # Weekly time series
  dates <- seq(as.Date("1970-01-01"), by = "week", length.out = n)
  
  # Temperature: seasonal pattern
  week_of_year <- as.numeric(format(dates, "%V"))
  tempr <- 65 + 20 * sin(2 * pi * week_of_year / 52) + rnorm(n, 0, 5)
  
  # Particulates: correlated with temperature (inversions) + trend
  part <- 50 + 0.3 * (tempr - 65) - 0.02 * (1:n) + rnorm(n, 0, 8)
  part <- pmax(20, part)  # Keep positive
  
  # Mortality: increases with temperature extremes and pollution
  temp_effect <- 0.5 * ((tempr - 65)^2) / 100
  pollution_effect <- 0.3 * (part - 50) / 10
  cmort <- 80 + temp_effect + pollution_effect + 
    2 * sin(2 * pi * week_of_year / 52) + rnorm(n, 0, 5)
  cmort <- pmax(60, cmort)
  
  # Return as time series objects
  list(
    cmort = ts(cmort, start = c(1970, 1), frequency = 52),
    tempr = ts(tempr, start = c(1970, 1), frequency = 52),
    part = ts(part, start = c(1970, 1), frequency = 52)
  )
}

generate_aids_data <- function() {
  # Generate monthly AIDS diagnoses and deaths (1987-1996)
  set.seed(789)
  n <- 120  # 10 years * 12 months
  dates <- seq(as.Date("1987-01-01"), by = "month", length.out = n)
  
  # Diagnoses: rise then fall with AZT introduction
  peak <- 60
  diag <- 500 + 800 * exp(-((1:n - peak)^2) / 500) + rnorm(n, 0, 50)
  diag <- pmax(100, diag)
  
  # Deaths: lag diagnoses by ~6-12 months
  death <- numeric(n)
  for(i in 13:n) {
    death[i] <- 0.6 * diag[i-6] + 0.3 * diag[i-12] + rnorm(1, 0, 30)
  }
  death <- pmax(50, death)
  death[1:12] <- NA  # Missing first year
  
  data.frame(
    month = dates,
    diag = round(diag),
    deaths = round(death)
  )
}
```

## Learning Objectives

By the end of this module, you should be able to:

1. Understand when and why to use VAR models over univariate ARIMA
2. Estimate VAR models and interpret system dynamics
3. Use impulse response functions (IRFs) to trace shock propagation
4. Apply variance decomposition to understand variable interactions
5. Recognize and test for structural breaks in time series relationships

## Topics

- Vector Autoregression (VAR)
- Impulse Response Functions
- Variance Decomposition
- Coefficient Dynamics & Structural Breaks

# Motivation: The Feedback Problem

## The Limitation of ARDL

Recall Autoregressive Distributed Lag models:

$$Y_t = \alpha_1 Y_{t-1} + \beta_1 X_{t-1} + e_t$$

. . .

**Problem**: This assumes $X$ "causes" $Y$, but $Y$ cannot influence $X$

. . .

**Real world**: Most economic variables have **bidirectional** relationships!

- GDP ↔ Unemployment
- Interest rates ↔ Inflation
- Pollution ↔ Temperature
- Diagnoses ↔ Deaths

## The VAR Solution

Allow **feedback** by modeling all variables as endogenous:

$$\begin{align}
Y_t &= \alpha_1 Y_{t-1} + \beta_1 X_{t-1} + e_{1t}\\
X_t &= \alpha_2 Y_{t-1} + \beta_2 X_{t-1} + e_{2t}
\end{align}$$

. . .

Now:
- Past $X$ predicts future $Y$ ✓
- Past $Y$ predicts future $X$ ✓
- Both variables evolve together as a **system**

## Historical Context

Chris Sims's [Macroeconomics and Reality](https://www.jstor.org/stable/pdf/1912017.pdf) (1980) revolutionized empirical macroeconomics.

. . .

**Key insight**: Let the data speak rather than imposing restrictive economic theory

- Traditional: Specify which variable is "exogenous"
- VAR: Treat all variables symmetrically, let lags capture relationships

. . .

**Impact**: VAR became the workhorse model for:
- Central bank forecasting
- Policy analysis
- Business cycle research

# VAR Mechanics

## The Challenge: Correlated Errors

Since we're estimating a system, errors might be correlated!

. . .

**Solution**: Decompose into orthogonal (independent) shocks

$$\begin{align} 
e_{1t} &= u_{1t} \\
e_{2t} &= \rho e_{1t} + u_{2t} = \rho u_{1t} + u_{2t}
\end{align}$$

where $u_{1t}$ and $u_{2t}$ are independent.

## Ordering and Identification

The full system becomes:

$$\begin{align}
Y_t &= \alpha_1 Y_{t-1} + \beta_1 X_{t-1} + u_{1t}\\
X_t &= \alpha_2 Y_{t-1} + \beta_2 X_{t-1} + \rho u_{1t} + u_{2t}
\end{align}$$

. . .

**Critical**: Variable ordering matters!

- Shock to $Y$ impacts both $Y$ and $X$ **immediately**
- Shock to $X$ only affects $Y$ next period (through lag)

. . .

**Interpretation**: Variable listed first is "more exogenous"

## Ordering Examples

**Monetary Policy VAR** (common ordering):
1. GDP → affects everything with delay
2. Inflation → central banks respond to it
3. Interest rate → policy tool (responds fastest)

. . .

**Environmental Health VAR**:
1. Pollution → external process
2. Temperature → responds to pollution/weather
3. Mortality → responds to both

. . .

**Rule of thumb**: More exogenous/slow-moving variables first

# Estimating VAR in R

## The `vars` Package

Main functions ([reference manual](https://cran.r-project.org/web/packages/vars/vars.pdf)):

```r
VAR()         # Estimate VAR model
VARselect()   # Choose optimal lag length
irf()         # Impulse response functions
fevd()        # Forecast error variance decomposition
predict()     # Forecasting
restrict()    # Restricted VAR (drop insignificant lags)
```

## VAR() Function

Key arguments:

- `y`: Dataset (must be `ts` object)
- `p`: Number of lags
- `type`: "none", "const", "trend", "both"
- `season`: Integer for seasonal dummies (e.g., 52 for weekly)
- `exogen`: Exogenous variables (no feedback)

. . .

```r
VAR(y = data,           # Time series object
    p = 2,              # 2 lags
    type = "const",     # Include constant
    season = NULL)      # No seasonal dummies
```

# Simulated VAR Example

## Generate Bivariate VAR Data {.smaller}

```{r}
#| code-fold: show
# True DGP: VAR(2) with feedback
N <- 200
y <- rep(0, N)
x <- rep(0, N)

# Correlated shocks (y shock affects both)
shocksy <- rnorm(N, sd = 0.2)
shocksx <- 0.5*shocksy + rnorm(N, sd = 0.2)

for(i in 3:N){
  # Y equation: depends on own lags and X lags
  y[i] <- 0.4*y[i-1] - 0.2*y[i-2] - 0.3*x[i-1] + 0.6*x[i-2] + shocksy[i]
  
  # X equation: depends on own lags and Y lags (feedback!)
  x[i] <- 0.5*y[i-1] - 0.5*y[i-2] + 0.5*x[i-1] + 0.4*x[i-2] + shocksx[i]
}

sim <- ts(cbind(y, x), start = 1, frequency = 1)
head(sim)
```

## Visualize the Data {visibility="uncounted" .smaller}

```{r}
#| code-fold: true
par(mfrow = c(1, 2))
plot(y, type = "l", main = "Series Y", ylab = "Y")
plot(x, type = "l", main = "Series X", ylab = "X")
par(mfrow = c(1, 1))
```

Both series show autocorrelation and appear stationary.

## Check Stationarity {visibility="uncounted"}

```{r results='hold'}
#| code-fold: show
adf.test(sim[,"y"])
adf.test(sim[,"x"])
```

Both reject unit root → proceed with VAR in levels.

## Examine Cross-Correlations {visibility="uncounted" .smaller}

::: {.panel-tabset}

### ACF
```{r}
#| code-fold: true
acf(sim, main = "Autocorrelation Function")
```

### PACF
```{r}
#| code-fold: true
pacf(sim, main = "Partial Autocorrelation Function")
```

:::

Off-diagonal ACF shows cross-correlation between series.

## Select Lag Length {visibility="uncounted"}

```{r}
#| code-fold: show
lagz <- VARselect(sim, lag.max = 8)
lagz$selection
```

AIC suggests 2 lags → matches true DGP!

## Estimate VAR(2) {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
# True model:
# y[i] <- 0.4*y[i-1] - 0.2*y[i-2] - 0.3*x[i-1] + 0.6*x[i-2] + shocksy[i]
# x[i] <- 0.5*y[i-1] - 0.5*y[i-2] + 0.5*x[i-1] + 0.4*x[i-2] + shocksx[i]

reg <- VAR(sim, p = 2, type = "none")  # No constant (matches DGP)
summary(reg)
```

## Interpreting VAR Output {visibility="uncounted"}

**Y equation**: Past values of both Y and X predict Y
- `y.l1` (0.39) ≈ true 0.4 ✓
- `x.l1` (-0.30) ≈ true -0.3 ✓

**X equation**: Past values of both Y and X predict X
- `y.l1` (0.48) ≈ true 0.5 ✓
- `x.l2` (0.39) ≈ true 0.4 ✓

. . .

**But how do we interpret system dynamics?**

→ Use **Impulse Response Functions**!

# Impulse Response Functions

## What are IRFs?

**Question**: If we shock variable $X$ today, how does variable $Y$ respond over time?

. . .

**IRF traces**:
- Immediate impact (period 0)
- Short-run dynamics (periods 1-5)
- Long-run effects (periods 10+)
- Speed of convergence back to equilibrium

. . .

**Why useful?**
- Individual coefficients hard to interpret
- IRFs show total accumulated effect
- Visualize shock propagation through system

## Computing IRFs

```r
irf(x,                    # Fitted VAR object
    impulse = "y",        # Which variable to shock
    response = "x",       # Which variable's response to track
    n.ahead = 10,         # Forecast horizon
    ortho = TRUE,         # Use orthogonal shocks
    cumulative = FALSE)   # Show levels (not cumulative)
```

## IRF: Shock to Y → Y Response {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
impulse <- irf(reg, impulse = "y", response = "y", n.ahead = 10)
plot(impulse, main = "Response of Y to Y Shock")
```

One-unit shock to Y → Y increases immediately, persists for ~5 periods.

## IRF: Shock to Y → X Response {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
impulse <- irf(reg, impulse = "y", response = "x", n.ahead = 10)
plot(impulse, main = "Response of X to Y Shock")
```

Shock to Y → X increases next period (feedback effect).

## IRF: Shock to X → Y Response {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
impulse <- irf(reg, impulse = "x", response = "y", n.ahead = 10)
plot(impulse, main = "Response of Y to X Shock")
```

Shock to X → Y responds negatively (negative coefficient in Y equation).

## IRF: Shock to X → X Response {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
impulse <- irf(reg, impulse = "x", response = "x", n.ahead = 10)
plot(impulse, main = "Response of X to X Shock")
```

Shock to X → X increases immediately, gradually decays.

## IRF Interpretation Guide

**Reading IRFs**:

1. **Point estimate** (solid line): Expected response path
2. **Confidence bands** (dashed): Uncertainty around estimate
3. **Zero crossing**: When effect becomes insignificant
4. **Persistence**: How long shock matters

. . .

**Key questions**:
- Does response become significant?
- How long does effect last?
- Does variable overshoot/oscillate?
- What's the cumulative effect?

# Variance Decomposition

## What is FEVD?

**Forecast Error Variance Decomposition** answers:

> "What % of forecast error variance in Y is due to shocks in X?"

. . .

**Interpretation**:
- High % → X is important for predicting Y
- Low % → Y mostly driven by its own shocks
- Changes over horizon → short vs long-run drivers differ

## FEVD for Simulated Data {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
plot(fevd(reg), main = "Variance Decomposition")
```

**Y decomposition**: Mostly own shocks (~80%), some X influence

**X decomposition**: Y shocks matter more (~40%) → strong feedback

# Forecasting with VAR

## VAR Forecasting

VAR forecasts all variables simultaneously:

```r
predict(var_model,      # Fitted VAR
        n.ahead = 20,   # Forecast horizon
        ci = 0.95)      # Confidence level
```

. . .

**Advantages over univariate**:
- Uses cross-variable information
- Consistent system forecasts
- Respects dynamic interactions

## Forecast Simulation {visibility="uncounted" .smaller}

::: {.panel-tabset}

### Y Forecast

```{r}
#| code-fold: true
pred <- predict(reg, n.ahead = 20)
plot(y, type = "l", xlim = c(1, N + 20), main = "Y Forecast")
lines((N+1):(N+20), pred$fcst$y[,1], col = "tomato", lwd = 2)
lines((N+1):(N+20), pred$fcst$y[,2], col = "tomato", lty = 2)
lines((N+1):(N+20), pred$fcst$y[,3], col = "tomato", lty = 2)
legend("topright", legend = c("Actual", "Forecast", "95% CI"),
       col = c("black", "tomato", "tomato"),
       lty = c(1, 1, 2), bty = "n")
```

### X Forecast

```{r}
#| code-fold: true
plot(x, type = "l", xlim = c(1, N + 20), main = "X Forecast")
lines((N+1):(N+20), pred$fcst$x[,1], col = "tomato", lwd = 2)
lines((N+1):(N+20), pred$fcst$x[,2], col = "tomato", lty = 2)
lines((N+1):(N+20), pred$fcst$x[,3], col = "tomato", lty = 2)
legend("topright", legend = c("Actual", "Forecast", "95% CI"),
       col = c("black", "tomato", "tomato"),
       lty = c(1, 1, 2), bty = "n")
```

### Fanchart

```{r}
#| code-fold: true
fanchart(pred, names = "x", main = "X Forecast with Uncertainty")
```

:::

# Application: LA Air Quality

## LA Air Quality Data {.smaller}

Weekly data from Los Angeles County (1970-1979):

- **Mortality** (`cmort`): Cardiovascular deaths
- **Temperature** (`tempr`): Weekly average temp (°F)
- **Particulates** (`part`): Air pollution measure

. . .

```{r}
#| code-fold: show
# Load or generate data
# Real data: from astsa package (cmort, tempr, part)
# Synthetic data:
la_data <- generate_la_air_quality()
cmort <- la_data$cmort
tempr <- la_data$tempr
part <- la_data$part
```

**Research question**: How do pollution and temperature affect mortality?

## Visualize LA Data {visibility="uncounted" .smaller}

::: {.panel-tabset}

### Mortality

```{r}
#| code-fold: true
plot(cmort, main = "Cardiovascular Mortality",
     ylab = "Deaths per 100k", xlab = "Year")
```

### Temperature

```{r}
#| code-fold: true
plot(tempr, main = "Average Temperature",
     ylab = "Degrees F", xlab = "Year")
```

### Pollution

```{r}
#| code-fold: true
plot(part, main = "Particulate Pollution",
     ylab = "Particulates", xlab = "Year")
```

:::

All show clear seasonality (summer peaks).

## Check Stationarity {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
adf.test(cmort)
adf.test(tempr)
adf.test(part)
```

All series are stationary → proceed with VAR.

## Variable Ordering Decision {visibility="uncounted"}

**Key choice**: What order for variables?

. . .

**Option 1**: Mortality → Temperature → Particulates
- Deaths most endogenous (respond to environment)
- Less plausible

. . .

**Option 2**: Particulates → Temperature → Mortality
- Pollution driven by external factors (cars, industry)
- Temperature partly responds to pollution (inversions)
- Mortality responds to both environmental factors
- **More plausible** ✓

## Prepare LA Data {visibility="uncounted"}

```{r}
#| code-fold: show
# Combine into multivariate time series
# Order: Particulates → Temperature → Mortality
la <- ts.union(part = part,
               tempr = tempr,
               cmort = cmort)

head(la, 10)
```

**Ordering assumptions**:
- Shock to particulates → affects particulates, temp, mortality
- Shock to temperature → affects temperature, mortality
- Shock to mortality → affects only mortality

## Examine LA Correlations {visibility="uncounted" .smaller}

::: {.panel-tabset}

### ACF
```{r}
#| code-fold: true
acf(la, lag.max = 52, main = "LA Air Quality ACF")
```

### PACF
```{r}
#| code-fold: true
pacf(la, lag.max = 52, main = "LA Air Quality PACF")
```

:::

Strong seasonality (52-week pattern) and cross-correlations visible.

## Select Lag Length {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
# Select lags accounting for seasonality
lagz <- VARselect(la,
                  lag.max = 26,   # Test up to half-year lags
                  season = 52)    # Include seasonal dummies
lagz$selection
```

AIC/FPE suggest 5 lags.

## Estimate LA VAR Model {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
# Estimate VAR(5) with trend and seasonal dummies
reg_la <- VAR(la,
              p = 5,            # 5 lags
              type = "both",    # Constant + trend
              season = 52)      # Weekly seasonal dummies

summary(reg_la)
```

## Interpreting LA Results {visibility="uncounted"}

**Many coefficients!** (5 lags × 3 variables × 3 equations = 45 params)

. . .

**Instead of interpreting individual coefficients**:
- Use IRFs to understand shock transmission
- Use FEVD to quantify variable importance
- Focus on economically meaningful questions

. . .

**Better questions**:
- If pollution spikes, how does mortality respond?
- How long do temperature effects last?
- What drives mortality variation: own history, pollution, or temperature?

## LA Impulse Responses {visibility="uncounted" .smaller}

View interactively in RStudio:

```{r eval=FALSE}
#| code-fold: show
impulse_la <- irf(reg_la, n.ahead = 20)
plot(impulse_la)
```

**Key findings** (from 9 IRF plots):
- Pollution shock → mortality increases after 2-3 weeks
- Temperature shock → immediate mortality response
- Effects persist 3-6 weeks before dissipating

## LA Variance Decomposition {visibility="uncounted" .smaller}

```{r eval=FALSE}
#| code-fold: show
fevd_la <- fevd(reg_la, n.ahead = 20)
plot(fevd_la)
```

**Mortality decomposition**:
- Short run: Mostly own shocks (~80%)
- Long run: Temperature and pollution matter more (~30%)
- Suggests environmental factors important but mortality has inertia

## Forecasting LA Mortality {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
# Forecast 26 weeks ahead (half year)
H <- 26
pred_la <- predict(reg_la, n.ahead = H)

# Create time sequence for forecast
xlimz <- c(min(time(cmort)), max(time(cmort)) + (H/52))
xz <- seq(max(time(cmort)) + (1/52), max(time(cmort)) + (H/52), 1/52)
```

## LA Forecasts: Mortality {visibility="uncounted" .smaller}

```{r}
#| code-fold: true
plot(cmort, xlim = xlimz, main = "Mortality Forecast",
     ylab = "Deaths", xlab = "Year")
lines(xz, pred_la$fcst$cmort[,1], col = "tomato", lwd = 2)
lines(xz, pred_la$fcst$cmort[,2], col = "tomato", lty = 2)
lines(xz, pred_la$fcst$cmort[,3], col = "tomato", lty = 2)
legend("topright", legend = c("Actual", "Forecast", "95% CI"),
       col = c("black", "tomato", "tomato"), lty = c(1, 1, 2), bty = "n")
```

## LA Forecasts: All Variables {visibility="uncounted" .smaller}

::: {.panel-tabset}

### Temperature
```{r}
#| code-fold: true
plot(tempr, xlim = xlimz, main = "Temperature Forecast")
lines(xz, pred_la$fcst$tempr[,1], col = "tomato", lwd = 2)
```

### Pollution
```{r}
#| code-fold: true
plot(part, xlim = xlimz, main = "Particulate Forecast")
lines(xz, pred_la$fcst$part[,1], col = "tomato", lwd = 2)
```

:::

Forecasts capture seasonal patterns and relationships.

# Application: AIDS Deaths

## AIDS Forecasting Challenge {.smaller}

Recall from Week 11: Can we beat univariate ARIMA using diagnoses?

. . .

**Previous approach**: ARDL with lagged diagnoses

**This week**: VAR allowing diagnoses ↔ deaths feedback

. . .

```{r}
#| code-fold: show
# Generate or load AIDS data
tmp3 <- generate_aids_data()

# Aggregate and prepare
tmp <- aggregate(list(death = tmp3$deaths, diag = tmp3$diag),
                 list(month = tmp3$month), sum, na.rm = TRUE)
tmp$death <- ifelse(tmp$death == 0, NA, tmp$death)
tmp$diag <- ifelse(tmp$diag == 0, NA, tmp$diag)
tmp <- tmp[year(tmp$month) >= 1987,]

# Split train/test at 1994-04-01
lim1 <- tmp$month <= ymd("1994-04-01") & !is.na(tmp$death)
```

## Baseline: ARIMA Forecast {visibility="uncounted" .smaller}

```{r echo=FALSE}
# Helper function for plotting
plot_all <- function(){
  plot(tmp$month, tmp$diag, type = "n",
       xlab = "Month", ylab = "Count",
       xlim = c(min(tmp$month), ymd("1999-01-01")),
       main = "AIDS Deaths: Forecast Comparison")
  lines(tmp$month[lim1], tmp$diag[lim1], col = "black", lwd = 2)
  lines(tmp$month[lim1], tmp$death[lim1], col = "dodgerblue", lwd = 2)
  abline(v = ymd("1994-04-15"), lty = 2)
  legend("topright", bty = "n", col = c("black", "dodgerblue"), lty = 1,
         legend = c("Diagnoses", "Deaths (train)"), lwd = 2)
}

# Fit ARIMA benchmark
aids_ts <- ts(tmp[lim1, c("diag", "death")],
              start = c(min(year(tmp$month[lim1]))),
              frequency = 12)

reg_arima <- auto.arima(aids_ts[,"death"])
H <- 24
pred_arima <- predict(reg_arima, n.ahead = H)

# Plot
plot_all()
lines(tmp$month[!lim1][1:H], pred_arima$pred,
      col = "tomato", lwd = 2)
lines(tmp$month[!lim1][1:H], tmp$death[!lim1][1:H],
      col = "mediumseagreen", lwd = 2)
legend("bottomright", bty = "n", lwd = 2,
       col = c("tomato", "mediumseagreen"), lty = 1,
       legend = c("ARIMA Forecast", "Actual Deaths"))
```

ARIMA misses the declining trend (doesn't use diagnoses info).

## AIDS VAR Setup {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
# Select optimal lag length
lagz_aids <- VARselect(aids_ts, lag.max = 36)
lagz_aids$selection
```

AIC suggests 26 lags, but this creates estimation problems (too many parameters).

**Better choice**: Use 12 lags (1 year) for stable estimation and interpretability.

## Estimate AIDS VAR {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
# VAR with reasonable lags to avoid overfitting
# Note: AIC suggested 26, but that's too many parameters for sample size
# Use 12 lags (1 year) as more stable choice
reg_var <- VAR(aids_ts, p = 12, type = "const")  # Constant only (no trend)

# Try to restrict to significant coefficients
reg_var_restricted <- tryCatch({
  restrict(reg_var, method = "ser")
}, error = function(e) {
  message("Note: Automatic restriction failed, using full model")
  reg_var
})

# Display results (with error handling)
tryCatch({
  summary(reg_var_restricted)
}, error = function(e) {
  message("Note: Summary failed, showing model structure instead")
  print(reg_var_restricted)
})
```

If restriction succeeds, model drops many insignificant lags for parsimony.

**Lag Selection Lesson**: AIC often suggests more lags than are practically useful. Consider:
- Sample size constraints (need 3-5 obs per parameter)
- Interpretability (12 lags = 1 year makes sense)
- Forecast stability (fewer lags often forecast better)

## AIDS VAR Forecast {visibility="uncounted" .smaller}

```{r}
#| code-fold: true
# Forecast using the (possibly restricted) model
pred_var <- predict(reg_var_restricted, n.ahead = H)

plot_all()
lines(tmp$month[!lim1][1:H], pred_var$fcst$death[,1],
      col = "tomato", lwd = 2)
lines(tmp$month[!lim1][1:H], tmp$death[!lim1][1:H],
      col = "mediumseagreen", lwd = 2)
legend("bottomright", bty = "n", lwd = 2,
       col = c("tomato", "mediumseagreen"), lty = 1,
       legend = c("VAR Forecast", "Actual Deaths"))
```

VAR captures the decline better!

## Forecast Accuracy Comparison {visibility="uncounted" .smaller}

```{r}
#| code-fold: show
# Mean Absolute Percentage Error
mape <- function(actual, forecast) {
  mean(abs((actual - forecast) / actual)) * 100
}

# Compare models
actual_deaths <- tmp$death[!lim1][1:H]
cbind(
  VAR = round(mape(actual_deaths, pred_var$fcst$death[,1]), 2),
  ARIMA = round(mape(actual_deaths, pred_arima$pred), 2),
  Average = round(mape(actual_deaths,
                       (pred_var$fcst$death[,1] + pred_arima$pred)/2), 2)
)
```

**VAR wins!** Even better: average both forecasts (forecast combination).

## Model Averaging {visibility="uncounted" .smaller}

```{r}
#| code-fold: true
plot_all()
lines(tmp$month[!lim1][1:H],
      (pred_arima$pred + pred_var$fcst$death[,1])/2,
      col = "tomato", lwd = 3)
lines(tmp$month[!lim1][1:H], pred_arima$pred,
      col = alpha("grey", 0.6), lwd = 2)
lines(tmp$month[!lim1][1:H], pred_var$fcst$death[,1],
      col = alpha("orchid", 0.6), lwd = 2)
lines(tmp$month[!lim1][1:H], tmp$death[!lim1][1:H],
      col = "mediumseagreen", lwd = 3)
legend("topright", bty = "n", lwd = 2,
       col = c("tomato", "grey", "orchid", "mediumseagreen"),
       legend = c("Average", "ARIMA", "VAR", "Actual"))
```

# Coefficient Dynamics

## The Stability Assumption

**So far**: We've assumed relationships are constant over time

$$Y_t = \beta X_t + e_t \quad \text{(same } \beta \text{ always)}$$

. . .

**Reality**: Economic relationships evolve!

- Weight vs Height (NBA player size evolution)
- Unemployment vs Inflation (Phillips Curve breakdown)
- Policy effectiveness (regime changes)

## Phillips Curve Example

<center>
<iframe src="https://fred.stlouisfed.org/graph/graph-landing.php?g=WtqU&width=960&height=500" scrolling="no" frameborder="0" style="width:960px; height:500px;" allowTransparency="true" loading="lazy" data-external="1"></iframe>
</center>

Relationship between unemployment and inflation not stable over time!

## Tools for Testing Stability

Two main approaches:

1. **Chow Break Test** (package: `strucchange`)
   - Test for breaks at known/suspected dates
   - Identify optimal break points

2. **Rolling Regression** (can use base R or packages like `roll` or `runner`)
   - Estimate model over moving windows
   - Visualize coefficient evolution

## Chow Break Test {.smaller}

**Idea**: Test if coefficients differ before vs after a break point

. . .

**Test statistic**:

$$F = \frac{(RSS_{pooled} - (RSS_1 + RSS_2))/k}{(RSS_1 + RSS_2)/(N_1 + N_2 - 2k)}$$

where:
- $RSS_{pooled}$: Full sample residual sum of squares
- $RSS_1$, $RSS_2$: Subsample RSS before/after break
- $k$: Number of regressors

## Simulated Break Example {visibility="uncounted"}

```{r}
#| code-fold: true
# Generate data with structural break at t=80
N <- 200
x <- rnorm(N)
t1 <- 1:N
t2 <- c(rep(0, 0.4*N), 1:(0.6*N))  # Kink at 40% point

# Y changes relationship after break
y <- 3*x + t1 - t2 + rnorm(N)

plot(t1, y, type = "l", main = "Series with Structural Break",
     xlab = "Time", ylab = "Y")
abline(v = 80, col = "red", lty = 2)
text(80, max(y), "Break?", pos = 4, col = "red")
```

## Test for Break at t=80 {visibility="uncounted"}

```{r}
#| code-fold: show
library(strucchange)
sctest(y ~ x, type = "Chow", point = 80)
```

Strong evidence of structural break! (p < 0.001)

## Finding the Break Point {visibility="uncounted" .smaller}

If you don't know when the break occurred:

```{r}
#| code-fold: show
# Test every potential break point (exclude ends)
break_stats <- sapply(20:180, function(i) {
  sctest(y ~ x, type = "Chow", point = i)$statistic
})
```

## Visualize Break Detection {visibility="uncounted"}

```{r}
#| code-fold: true
plot(20:180, break_stats, type = "b", pch = 19,
     xlab = "Potential Break Point", ylab = "Chow Statistic",
     main = "Identifying Structural Break")
abline(v = 80, col = "red", lty = 2)
text(80, max(break_stats), "True Break", pos = 4, col = "red")
```

Peak at t=80 identifies the true break point!

## Using Break Tests

**Practical applications**:

1. **Sample selection**: Stop analysis before known policy change
2. **Break modeling**: Include dummies/interactions for regime shifts
3. **Model validation**: Test if relationship stable over evaluation period

. . .

**Example uses**:
- Pre vs post-Great Recession
- Policy implementation dates
- Technology adoption thresholds

# Rolling Regression

## Rolling Regression Concept

Estimate model repeatedly over moving windows:

**Shifting window**: Fixed width, moves forward
```
[Window 1: t=1 to 25]
    [Window 2: t=2 to 26]
        [Window 3: t=3 to 27]
```

**Expanding window**: Grows over time
```
[Window 1: t=1 to 25]
[Window 2: t=1 to 26]
[Window 3: t=1 to 27]
```

## Rolling Regression in R

**Simple base R implementation** (no packages needed):

```{r eval=FALSE}
#| code-fold: show
# Rolling coefficient function
rolling_coef <- function(y, x, width, expanding = FALSE) {
  n <- length(y)
  coefs <- matrix(NA, nrow = n - width + 1, ncol = 2)
  
  for(i in width:n) {
    if(expanding) {
      idx <- 1:i              # Expanding: use all data from start
    } else {
      idx <- (i-width+1):i    # Shifting: use last 'width' obs
    }
    fit <- lm(y[idx] ~ x[idx])
    coefs[i - width + 1, ] <- coef(fit)
  }
  return(coefs)
}

# Use it
coefficients <- rolling_coef(y, x, width = 25, expanding = FALSE)
```

**Alternatives**: Packages like `roll`, `runner`, or `slider` offer more features.

## Rolling Regression Example {visibility="uncounted"}

```{r}
#| code-fold: show
# Simple rolling regression function (no extra packages needed)
rolling_coef <- function(y, x, width, expanding = FALSE) {
  n <- length(y)
  coefs <- matrix(NA, nrow = n - width + 1, ncol = 2)
  colnames(coefs) <- c("(Intercept)", "x")
  
  for(i in width:n) {
    if(expanding) {
      # Expanding window: use all data from start to i
      idx <- 1:i
    } else {
      # Shifting window: use last 'width' observations
      idx <- (i - width + 1):i
    }
    fit <- lm(y[idx] ~ x[idx])
    coefs[i - width + 1, ] <- coef(fit)
  }
  return(coefs)
}

# Estimate with both window types
r_shift <- rolling_coef(y, t1, width = 25, expanding = FALSE)
r_expand <- rolling_coef(y, t1, width = 25, expanding = TRUE)
```

## Visualize Coefficient Evolution {visibility="uncounted"}

```{r}
#| code-fold: true
par(mfrow = c(1, 2))

# Shifting window - coefficient on t1 (second column after intercept)
plot(r_shift[,"x"], type = "b", pch = 19,
     main = "Shifting Window", ylim = c(-0.1, 1.1),
     xlab = "Time", ylab = "Coefficient on t1")
abline(h = c(0, 1), lty = 2, col = "grey")

# Expanding window
plot(r_expand[,"x"], type = "b", pch = 19,
     main = "Expanding Window", ylim = c(-0.1, 1.1),
     xlab = "Time", ylab = "Coefficient on t1")
abline(h = c(0, 1), lty = 2, col = "grey")

par(mfrow = c(1, 1))
```

Shifting window detects break faster; expanding window shows gradual adjustment.

## When to Use Each Tool

| Tool | Best for | Limitation |
|------|----------|------------|
| **Chow Test** | Known break dates; Single break | Requires break point specification |
| **Rolling Regression** | Unknown breaks; Gradual changes | Computationally intensive; harder to test |

. . .

**Best practice**: Use both!
1. Rolling regression to visualize dynamics
2. Chow test to formally test suspected breaks

## VAR and Structural Breaks

**Important**: VAR coefficients can also break!

- Monetary policy regime changes
- Technological shocks
- Financial crises

. . .

**Solutions**:
1. Split sample around known breaks
2. Include structural break dummies
3. Use Time-Varying Parameter VAR (advanced)

# Summary and Best Practices

## When to Use VAR

**Use VAR when**:
- Multiple endogenous variables with feedback
- Interested in system dynamics (not just Y on X)
- Want to trace shock transmission
- Need coherent multivariate forecasts

. . .

**Don't use VAR when**:
- Clear exogenous/endogenous distinction
- Small sample (VAR has many parameters!)
- Variables have very different frequencies

## VAR Workflow

1. **Check stationarity** → Transform if needed
2. **Choose variable ordering** → Use economic reasoning
3. **Select lag length** → Use VARselect()
4. **Estimate model** → VAR()
5. **Simplify** → Consider restrict() for parsimony
6. **Interpret dynamics** → Use irf() and fevd()
7. **Validate** → Check residuals, test for breaks
8. **Forecast** → predict()

## Common Pitfalls

❌ **Arbitrary ordering** → Affects contemporaneous effects

❌ **Ignoring stationarity** → Spurious dynamics

❌ **Over-interpretation of individual coefficients** → Use IRFs!

❌ **Too many lags** → Overfitting, poor forecasts, numerical instability (need 3-5 obs per parameter)

❌ **Assuming stability** → Test for structural breaks

## Key Takeaways

1. **VAR captures feedback** between variables
2. **Ordering matters** for contemporaneous effects
3. **IRFs reveal dynamics** better than raw coefficients
4. **FEVD quantifies** relative importance
5. **Structural breaks** can invalidate VAR relationships
6. **Rolling regression** helps detect instability

## Next Steps

::: {.nonincremental}

**Week 13 (11/30)**:

- Spatial Autocorrelation
- Project consultations

**Week 14 (12/7)**:

- Final Project Presentations

:::

**Today**: Please submit project data if you haven't already!
